---
title: 'Remote Sensing Lab 3: Satellite data resolution
  in R'
author: "Victor Gutierrez (victorhugo@temple.edu)"
output:
  html_document:
    df_print: paged
  pdf_document: default
  word_document: default
always_allow_html: yes
---

## Lab due
September 21 2021

## Goals
1. To learn basic techniques for image cropping, zooming in, and pixel aggregation/disaggregation.
2.	To understand differences in spatial, temporal, spectral and radiometric resolutions between optical satellites
3.	To understand trade-offs between satellite data under different spatial, temporal, radiometric and spectral resolutions.

## Total score
The lab counts for up to 4 points towards the final grade of the course.

Have fun!

```{r include = FALSE}
wd="/Users/tug61163/Library/CloudStorage/OneDrive-TempleUniversity/Courses/IntroRemoteSensing/2023Fall/Class3/LabMaterials_Class3_2023"
setwd(wd)
library(knitr)    # For knitting document and include_graphics function
#library(kableExtra)
library(ggplot2)  # For plotting
library(png)      # For grabbing the dimensions of png files
library(kableExtra)
fig1path=paste(wd, "Fig1.png", sep="/")
#fig2path=paste(wd, "Fig2.png", sep="/")
#fig3path=paste(wd, "Fig3.png", sep="/")
#fig4path=paste(wd, "Fig2NAIP.png", sep="/")
#fig5path=paste(wd, "Fig3NAIP.png", sep="/")
#fig6path=paste(wd, "Fig4.png", sep="/")

#fig1 = readPNG(fig1path)
#fig2 = readPNG(fig2path)
#fig3 = readPNG(fig3path)
#fig4 = readPNG(fig4path)
#fig5 = readPNG(fig5path)
#fig6 = readPNG(fig6path)
```

## Part A. Downloading images

We will reproduce the steps for data downloading described in lab 2 but this time, we will download other images. You will also need to copy the compressed (.tar or .tar.gz).  For detailed downloading instructions, please refer to the instructions for lab 2.

1. Go to Earth explorer (http://earthexplorer.usgs.gov). Click on the login link and enter the credentials that you used in class to open an account.

2.	Click on the “Search Criteria” tab, then type “Philadelphia” in the "Feature Name", select Pennsylvania in the "State"  and then click "Show". Click on the feature to show the location in the map view. This willshow a pin in the selected location in the map (for more details about this step, please refer to the lab from the previous class).

Zoom to the pin. Then locate Fairmount Park in the map and drag the pin with your mouse towards  an edge of the park. Click on  other edges until to produce a polygon that covers approximately the area of the park (see figure below).
```{r include = TRUE, echo=FALSE}
# echo ommits the printing of anything within the chunk 
include_graphics(fig1path)
```

3. Click on the “Data Sets” tab at the top of the search panel. Then on the drop down menu select Aerial Imagery/NAIP and checkmark the option NAIP .

4.	Click on the results tab and locate the NAIP image with an acaquisition date of 2022-05-10. If you click on the "Show Footprint" button, you will see in the map that this is the scene that covers most of the park. 

5.	. Click on the “Download Options”button" at the bottom of the entity (see instructions for lab 2) and download the "Full Resolution" image. The image should download in the default downloads folder in your computer as a compressed file with the name m_3907507_nw_18_060_20220510.ZIP.zip . Your commputer may decompress this image automatically depending on the configuration.

6. Move the compressed file into the working folder that you will specify as setwd() in R as shown in Part B. Also copy the compressed Landsat image that you used for lab 2 and paste it in the working folder. If you don't have access to this image anymore, you can follow the steps described in lab 2 to download it.

## Part B. Data import resampling, alignment and pixel aggregation/disaggregation
Load required libraries.
```{r include = TRUE, message=F, warning=F, eval=FALSE}
library(terra)
#library(rgdal)
#library(RStoolbox)
```

1. Set working directory: Change the path below to the folder where you stored your images in part A. This will be your working directory for this If you are using a windows machine, make sure you use forward slash instead of backward.
```{r include = TRUE, message=F, warning=F, eval=FALSE}
wd="/Users/tug61163/Library/CloudStorage/OneDrive-TempleUniversity/Courses/IntroRemoteSensing/2023Fall/Class3/LabMaterials_Class3_2023"
setwd(wd)
```

2. Check the names of the files in your working directory. Then reproduce the code from step 3 in lab 2 to decompress the image and open it in R as a raster stack.
```{r include = TRUE, message=F, warning=F, eval=FALSE}
untar("LC08_L2SP_014032_20230410_20230420_02_T1.tar")
tifs <- list.files('.', pattern='.TIF')
tifs # lists all the files in the working folder with the extension .TIF
tifBandNames=tifs[c(1, 4:9)] # enter the index position for the bands of interest (bands 1 through 7)
L8=rast(tifBandNames)
names(L8)=tifBandNames
names(L8)
L8
```

3. Check if the naip image was downloaded as a zip file or if it was already decompressed as a .tif file (this depends on the configuration of your computer). Copy and paste the name of the naip image below to decompress (if necessary) and then open it as a stack in R.
```{r include = TRUE, message=F, warning=F, eval=FALSE}
unzip("m_3907507_nw_18_060_20220510.ZIP")
naip=rast("m_3907507_nw_18_060_20220510.tif") # The tif file for NAIP has all the bands already stacked
```

4. Compare the spatial specifications of both images and then plot them as RGB composites 

```{r include = TRUE, message=F, warning=F, eval=FALSE}
L8 # Provides the spatial information associated with the downloaded image.
naip # Provides the spatial information associated with the downloaded image.
plotRGB(L8, r = 4, g = 3, b = 2,  stretch = "lin")
plotRGB(naip, r = 1, g = 2, b = 3, stretch = "lin")
```

5.Resize the Landsat image to the extent of the NAIP image.
```{r include = TRUE, message=F, warning=F, eval=FALSE}
L8rsz=crop(L8, naip)
plotRGB(L8rsz, r = 4, g = 3, b = 2,  stretch = "lin")
```

6. Aggregate the Landsat image so that the pixel size is approximately equal to the minimum resolution available for MODIS images (~232 m).
```{r include = TRUE, message=F, warning=F, eval=FALSE}
L8agg=aggregate(L8rsz, fact=8)
```

8. Plot the NAIP image along with the Landsat image after and before aggregation for the area covered by the extent of the NAIP image. Save the images as pdf. Use the drawExtent() function and the ext argument in the ggRGB function to zoom in into different areas of interest. Modifiy the stretch as needed to improve contrast. You can see stretch options by running ?plotRGB
```{r include = TRUE, message=F, warning=F, eval=FALSE}
plotRGB(L8agg, r = 4, g = 3, b = 2, stretch = "hist") # MODIS resolution (simulated)
plotRGB(L8rsz, r = 4, g = 3, b = 2, stretch = "hist") # Landsat resolution
plotRGB(naip, r = 1, g = 2, b = 3, stretch = "hist") # NAIP resolution
```

You can also plot the images with a different but equivalent band combination
```{r include = TRUE, message=F, warning=F, eval=FALSE}
plotRGB(L8agg, r = 5, g = 4, b = 3, stretch = "hist") # MODIS resolution (simulated)
plotRGB(L8rsz, r = 5, g = 4, b = 3, stretch = "hist") # Landsat resolution
plotRGB(naip, r = 4, g = 3, b = 2, stretch = "hist") # NAIP resolution
```

9. Save the three graphs as a single pdf file
```{r include = TRUE, message=F, warning=F, eval=FALSE}
pdf("VGutierrez_Lab3.pdf")
  plotRGB(L8agg, r = 5, g = 4, b = 3, stretch = "hist") # MODIS resolution (simulated)
  plotRGB(L8rsz, r = 5, g = 4, b = 3, stretch = "hist") # Landsat resolution
  plotRGB(naip, r = 4, g = 3, b = 2, stretch = "hist") # NAIP resolution
dev.off()
```

10. Save the images as raster files
```{r include = TRUE, message=F, warning=F, eval=FALSE}
writeRaster(L8agg, filename="MODISres.tif")
writeRaster(L8rsz, filename="LANDSATres.tif")
writeRaster(naip, filename="NAIPres.tif")
```

11. Open the three rasters in QGIS by clicking in the option "Layer" in the main menu, and then on "Add Layer" > "Add Raster Layer...". Browse to the working directory and select the three files saved with the names specified in step 10 and then click on "add".

12. Right click on each image and then select "properties. Change the band numbers to display as RGB, using a natural color band combination (the same as the one used in first set of RGB plots displayedin step 8). 

Zoom in out of different locations with all three images and interpret the features that are displayed..

Change again the band numbers to match the combination used for the second set of RGB plots displayed in step 8.

Zoom in out of different locations with all three images and interpret the features that are displayed. Which band combination is more helpful?

## Lab deliverables

Please provide the information requested in the "Lab3Deliverables.docx" file, available in module 3 in canvas.