---
title: 'Applied Machine Learning for Spatial Analysis. Lab Week 11: Supervised Regression'
author: "Victor Gutierrez (victorhugo@temple.edu)"
output:
  html_document:
    df_print: paged
  pdf_document: default
  word_document: default
---

```{r include = FALSE}
# Load screenshots
wd="/Users/tug61163/Library/CloudStorage/OneDrive-TempleUniversity/Repositories/Gitrepo/MALESA_Code"
setwd(wd)
library(knitr)    # For knitting document and include_graphics function
library(ggplot2)  # For plotting
library(png)      # For grabbing the dimensions of png files

fig1path=paste(wd, "Fig1.png", sep="/")
#fig2path=paste(wd, "Sel2Edit.png", sep="/")
#fig3path=paste(wd, "Triangle.png", sep="/")
#fig4path=paste(wd, "Fig4.png", sep="/")
#fig1 = readPNG(fig1path)
#fig2 = readPNG(fig2path)
#fig3 = readPNG(fig3path)
#fig4 = readPNG(fig4path)
```

## Goals
To learn:
1. How to fit and interpret a supervised regression using tree based methods and 
2. How to use the fitted model to produce spatial predictions.

## Lab instructions
1.	Launch R Studio and open a new R script: File/ New File/ R Script. Then save it as a new file: File/ Save As…
2.	Read the instructions below step by step. Copy and paste each chunk of code at a time in your R script. Select the code with your mouse or shift/arrow keys and then run it by pressing the keys control-enter simultaneously.

## Lab overview
For this lab, we are going to use data from forest inventories predicting aboveground biomass for the year 2019 in the state of NY. Data were obtained from the US Forest Inventory Data: https://www.fs.usda.gov/research/products/dataandtools/forestinventorydata


Have fun!

### 1. System setup
```{r include = TRUE, message=F, warning=F, eval=FALSE}
wd=("/Users/tug61163/Library/CloudStorage/OneDrive-TempleUniversity/Courses/MALESA/2024/Week11_SuperReg")
setwd(wd)
library(terra)
library(sf)
library(ggplot2)
library(corrplot)
library(DataExplorer)
library(caret)
library(summarytools)
library(randomForest)
library(gbm)
library(e1071)
library(xgboost)
library(purrr)
library(geodata)
library(leaflet)
library(tidycensus)
library(tigris)
library(dplyr)
library(prism)
library(iml)
prism_set_dl_dir(wd)
```

#### Load and format sampled datafiles

```{r include = TRUE, message=F, warning=F, eval=FALSE}
dataset=read.csv("plots.csv")
names(dataset)
```

Subset the data for a given year and then convert it into a spatial object
```{r include = TRUE, message=F, warning=F, eval=FALSE}
dataset=dataset[dataset$yr==2019,] # select data from 2019 only
summary(dataset$yr)
SpatData=st_as_sf(dataset, coords=c("long", "lat"), crs=st_crs("EPSG:4326"))
plot(SpatData["agb"], type="p", pch=4)
```

Let's explore the distribution of plot locations using leaflet
```{r include = TRUE, message=F, warning=F, eval=FALSE}
datasetPrj=st_transform(SpatData, "EPSG:4326")

# Extract coordinates from the point sf object
coords <- st_coordinates(datasetPrj)

# Add the coordinates back as columns in the sf object
datasetPrj$lon <- coords[, 1]  # Longitude
datasetPrj$lat <- coords[, 2]  # Latitude

# Now create the leaflet map using the extracted coordinates
leaflet() %>%
  addProviderTiles(providers$Esri.WorldImagery) %>%
  
  # Add the point layer using the preprocessed longitude and latitude
  addCircleMarkers(lng = ~lon, lat = ~lat,  # Use the new lon and lat columns
                   data = datasetPrj, 
                   color = "red", radius = 1, popup = ~as.character(agb)) %>%
  
  # Set the view of the map
  setView(lng = -76, lat = 43, zoom = 7)
```

Download and format input spatial variables
```{r include = TRUE, message=F, warning=F, eval=FALSE}
### Boundaries of NY State as the study area
options(tigris_class = "sf")
nyBound <- states(cb = TRUE, year = 2022) %>%
  filter(NAME == "New York")

### Topography
elev1=elevation_3s(-72, 42, wd)
elev2=elevation_3s(-77, 42, wd)
elev3=elevation_3s(-72, 47, wd)
elev=merge(elev1, elev2, elev3)
plot(elev)
#writeRaster(elev, "elev.tif")

# Project study area and sampled data using elevation as the reference
crs(nyBound)==crs(elev)
nyBound=terra::project(vect(nyBound), elev)
plot(elev)
plot(nyBound, add=T)

# Crop elevation to th eztent of NYC
elev=crop(elev, nyBound)
plot(elev)
plot(nyBound, add=T)

crs(elev)==crs(SpatData)
SpatDataPrj=terra::project(vect(SpatData), elev)
plot(elev)
plot(SpatDataPrj, add=T)
rm(SpatData, elev2)
writeVector(SpatDataPrj, "SpatDataPrj", filetype="ESRI Shapefile")

# Derive terrain variables from the elevation map
tervar=terrain(elev, v=c("slope", "aspect", "roughness", "TRI", "TPI", "flowdir"), filename="tervar.tif")
tervar=c(elev, tervar)
names(tervar)=c("elevation", "slope", "aspect", "roughness", "TRI", "TPI", "flowdir")
terra::writeRaster(tervar, "tervar1.tif", overwrite=T)
plot(tervar)
rm(elev)

### Canopy height from https://glad.umd.edu/dataset/gedi

# Copy this link and paste it in your browser to download the file: https://glad.geog.umd.edu/Potapov/Forest_height_2019/Forest_height_2019_NAM.tif

# Open, project and resize file
ch=rast("CH.tif")
plot(ch)

crs(ch)==crs(tervar)
ch=project(ch, tervar)
plot(ch)
plot(nyBound, add=T)
writeRaster(ch, "chprj.tif")

              
### Tree cover for from here: https://www.mrlc.gov/data?f%5B0%5D=category%3ATree%20Canopy 
# Copy the following link and paste it in your browser: https://s3-us-west-2.amazonaws.com/mrlc/nlcd_tcc_CONUS_2019_v2021-4.zip
unzip("nlcd_tcc_CONUS_2019_v2021-4.zip")
TC=rast("nlcd_tcc_conus_2019_v2021-4.tif")
crs(TC)==crs(tervar)
TC=terra::project(TC, tervar, method="near")
plot(TC)
plot(nyBound, add=T)

# We need to change the layer from categorical to numerical to make the file compatible with modeling
TC=as.numeric(TC)
TC[TC==255]<-NA
plot(TC)

writeRaster(TC, "TCprj.tif")
#rm(LC, tree, imp)
#LCdata=rast("LCdata.tif")
#LCdata=terra::project(LCdata[[1]], elev, method="near")

# Climate data
get_prism_annual(type="tmin", years = 2019, keepZip = TRUE, keep_pre81_months = FALSE)
get_prism_annual(type="tmax", years = 2019, keepZip = TRUE, keep_pre81_months = FALSE)
get_prism_annual(type="tmean", years = 2019, keepZip = TRUE, keep_pre81_months = FALSE)
get_prism_annual(type="tdmean", years = 2019, keepZip = TRUE, keep_pre81_months = FALSE)
get_prism_annual(type="vpdmin", years = 2019, keepZip = TRUE, keep_pre81_months = FALSE)
get_prism_annual(type="vpdmax", years = 2019, keepZip = TRUE, keep_pre81_months = FALSE)
get_prism_annual(type="ppt", years = 2019, keepZip = TRUE, keep_pre81_months = FALSE)

prisms <- list.files('.', pattern='_stable_4kmM3_2019_bil')

# Go to subfolders and reproject each dataset
for (i in seq(1,length(prisms), by=2)){
  prisrast=rast(paste0(wd, "/", prisms[i], "/", prisms[i], ".bil"))
  prisrast=project(prisrast, tervar)
  if(i==1){pristack=prisrast} else {pristack=c(pristack, prisrast)}
}
names(pristack)=substr(prisms[seq(1,length(prisms), by=2)], start=7, stop=10)

# shorten names for prism
names(pristack)= gsub("PRISM_|_stable_4kmM3_2019_bil", "", names(pristack))
plot(pristack)
writeRaster(pristack, "pristack1.tif", overwrite=T)
```

Stack all data in a single raster object
```{r include = TRUE, message=F, warning=F, eval=FALSE}
# Load files (If they are not already in the environment)
tervar=rast("tervar1.tif")
pristack=rast("pristack1.tif")
ch=rast("chprj.tif")
TC=rast("TCprj.tif")
plot(TC)

# Convert background pixels to (255) in tree cover map
tree[tree<0]=NA 
tree[tree>100]=NA 

datastack=c(pristack, tervar, ch, TC)
names(datastack)=c(names(pristack), names(tervar), "ch", "tc")

# remove out of bound pixels
nyBoundRast=terra::rasterize(nyBound, datastack)
datastack=datastack*nyBoundRast
plot(datastack)

writeRaster(datastack, "datastack.tif")
#rm(elev,tree, pristack, tervar)
```

##### Extract data into database
In order to expedite the running of the algorithms, I will subset the datasets before continuing.
```{r include = TRUE, message=F, warning=F, eval=FALSE}
# #SpatDataPrj=vect(paste0(wd, "/SpatDataPrj/SpatDataPrj.shp"))
# #datastack =rast("datastack.tif")
# names(datastack)
# plot(datastack[[15]])
# plot(SpatDataPrj, add=T)
# ext=terra::draw()
# 
# # For some reason is not allowing me to crop the whole dataset at once
# datastack=crop(datastack, ext)
# SpatDataPrj=terra::crop(SpatDataPrj, ext)
# 
# plot(datastack[[15]])
# plot(SpatDataPrj, add=T)

# Make sure the names in the datastack are properly labeled and that the projection matches with 
names(datastack)
crs(datastack)==crs(SpatDataPrj)
refdata=extract(datastack, SpatDataPrj)
nrow(refdata)
nrow(SpatDataPrj)
refdata=cbind(SpatDataPrj$agb, refdata)

# Make sure that the column names have the proper name
names(refdata)
names(refdata)[1]="agb"
save(refdata,file="refdata.RData")

refdatasub=refdata[,-2] # remove column representing ID
```

#### Explore the data and select variables
```{r include = TRUE, message=F, warning=F, eval=FALSE}
refdata %>%
    create_report(
        output_file = paste("Report", format(Sys.time(), "%Y-%m-%d %H:%M:%S %Z"), sep=" - "),
        report_title = "EDA Report - Plant communities",
        y = "agb"
    )
```
Based on the results of my EDA,these are the variables that I will select
```{r include = TRUE, message=F, warning=F, eval=FALSE}
response <- "agb"
selvar=c("ppt", "tmin", "vpdmax", "elevation", "slope", "ch", "tc")
```

#### Clean the data
There are missing observations that we have to remove
```{r include = TRUE, message=F, warning=F, eval=FALSE}
refdatasub=refdata[,-2] # remove column representing ID
refdatasub=subset(refdatasub, complete.cases(refdatasub)) # remove missing data
```

#### Fit different ML methods
WARNING 1: The most important task for the succesful run of the different algorithms is to make sure that input data is in the format accepted for data arguments.

##### Random forest
```{r include = TRUE, message=F, warning=F, eval=FALSE}
eq=as.formula((paste0(response, "~", (paste(selvar, collapse = " + ")))))
rf=randomForest(eq, data=refdatasub, ntree=1000)
rf
rf$importance
varImpPlot(rf)
```

##### Extreme boosting machine via xgboost
https://xgboost.readthedocs.io/en/stable/tutorials/model.html 
```{r include = TRUE, message=F, warning=F, eval=FALSE}
refdataX=refdatasub[c(selvar)]

#sampdataX_scaled <- scale(sampdataX)
agb=refdatasub$agb

dtrain <- xgb.DMatrix(data = as.matrix(refdataX), label = agb, missing = NA)

# Set parameters for the model
param <- list(
  objective = "reg:squarederror",  # regression squared loss
  eta = 0.1,  # learning rate
  max_depth = 6, # maximum depth of the tree
  eval_metric = "rmse"  # Error rate
)

# Train the model using xgb.train()
xgb_model <- xgb.train(params = param, data = dtrain, nrounds = 500, nthread = 6)

# Variable importance
importance <- xgb.importance(feature_names = colnames(refdataX), model = xgb_model)
head(importance)
xgb.plot.importance(importance_matrix = importance)
```

##### Support Vector Machines
```{r include = TRUE, message=F, warning=F, eval=FALSE}

# You should always normalize your data before doing SVM and kNN!! 
# From lab 5:
zscore=function(x){
  zval=(x-mean(x, na.rm=TRUE))/sd(x, na.rm=TRUE)
  return(zval)
}
refdataZ=data.frame(apply(refdataX, 2, FUN=zscore))
refdataZ=cbind(agb,refdataZ)
names(refdataZ)
str(refdataZ)
summary(refdataZ)

sv = svm(agb~., data = refdataZ, type = 'eps-regression', kernel = "sigmoid", cost = 10, scale = FALSE)
print(sv)

# I will clean up the graphical settings to make sure the plot below runs smoothly

while (!is.null(dev.list()))  dev.off()
dev.new()

# Plotting the SVM regression
if("elevation" %in% names(refdataZ)) {
  plot(refdataZ$ch, refdataZ$agb, main = "SVM Regression Plot",
       xlab = "Elevation", ylab = "AGB")
  points(refdataZ$elevation, predict(sv, refdataZ), col = "blue", pch = 16)
} else {
  print("Column 'elevation' not found in refdataZ")
}

# Blue line indicates the regression line or surface that the SVM model learned

# Fit Support Vector Machine model to data set
svtune <- tune(svm, agb~., data = refdataZ, kernel = "sigmoid",
               ranges = list(cost = c(0.001, 0.01, 0.1, 1, 5, 10, 100)))
# extract the best model
bestmod <- svtune$best.model
plot(svtune, main = "SVM Tuning Results")

# Get predictions from the best model
predictions <- predict(bestmod, refdataZ)

# Plot actual vs. predicted values
plot(refdataZ$agb, predictions, 
     xlab = "Actual AGB", ylab = "Predicted AGB", 
     main = "Actual vs. Predicted AGB")
abline(0, 1, col = "red")  # Add a 45-degree line for reference

```

#### Evaluate marginal effects

##### For random forest
```{r include = TRUE, message=F, warning=F, eval=FALSE}
### For one predictor

# Define predictor object for the model
predictor_rf <- Predictor$new(rf, data = refdataX, y = agb)

# Plot marginal effects for a specific feature
effect <- FeatureEffect$new(predictor_rf, feature = "ch", method = "pdp")
plot(effect)

effect2 <- FeatureEffect$new(predictor_rf, feature = "tc", method = "pdp")
plot(effect2)

effect3 <- FeatureEffect$new(predictor_rf, feature = "slope", method = "pdp")
plot(effect3)

effect4 <- FeatureEffect$new(predictor_rf, feature = "tmin", method = "pdp")
plot(effect4)

effect5 <- FeatureEffect$new(predictor_rf, feature = "ppt", method = "pdp")
plot(effect5)

### For two predictors
pdp_two_features_rf <- FeatureEffect$new(predictor_rf, 
                                         feature = c("slope", "tmin"), 
                                         method = "pdp")

# Plot the 2D PDP
plot(pdp_two_features_rf)
```

##### For xgboost
```{r include = TRUE, message=F, warning=F, eval=FALSE}
# Prepare the Predictor object from iml package
predictor_xgb <- Predictor$new(
  xgb_model, 
  data = as.data.frame(refdataX), 
  y = labels, 
  predict.function = function(model, newdata) {
    as.numeric(predict(model, as.matrix(newdata)))  # Ensure single-column output
  }
)

# Create a partial dependence plot for a single feature
pdp_iml <- FeatureEffect$new(predictor_xgb, feature = "tmin", method = "pdp")
plot(pdp_iml)

### For two predictors
# Define train_data if not already defined
train_data <- as.data.frame(refdataX)

# Assuming `labels` contains the target variable
predictor_xgb <- Predictor$new(
  xgb_model, 
  data = train_data, 
  y = labels, 
  predict.function = function(model, newdata) {
    as.numeric(predict(model, as.matrix(newdata)))
  }
)

# Generate a partial dependence plot for two features
pdp_iml_two <- FeatureEffect$new(
  predictor_xgb, 
  feature = c("slope", "elevation"),  # Specify the two features here
  method = "pdp"
)

# Plot the PDP for two features
plot(pdp_iml_two)
```


#### Produce predictions and save them
Let's first extract the data from the raster to a dataframe
```{r include = TRUE, message=F, warning=F, eval=FALSE}
datastackdf=datastack[[selvar]] # select from the stack a subset of bands corresponding to the variables selected for modeling
datastackdf = as.matrix(datastackdf, mode="numeric")
datastackdf=subset(datastackdf, complete.cases(datastackdf))
```

##### Random forest
```{r include = TRUE, message=F, warning=F, eval=FALSE}
rfpred=terra::predict(datastack, rf)
plot(rfpred)
writeRaster(rfpred, "rfrast.tif")
```

##### XGBoost
```{r include = TRUE, message=F, warning=F, eval=FALSE}
# Make sure that colnames are the same in raster and dataframe and correct if needed
colnames(datastackdf)==colnames(refdataX)

xgbpred <- stats::predict(xgb_model, datastackdf)
xgrast=sum(datastack)
xgrast[!is.na(xgrast)] <- 1
xgrast[!is.na(xgrast)] <- xgbpred
plot(xgrast)
writeRaster(xgrast, "xgbrast.tif")
```

##### SVM
```{r include = TRUE, message=F, warning=F, eval=FALSE}
# Standardize values in the raster stack so that the format is the same as the data used for calibrating the algorithm
datastackdfz=apply(datastackdf, 2, FUN=zscore)
colnames(datastackdfz)==colnames(refdataZ)[2:ncol(refdataZ)]
svm.pred <- predict(sv, datastackdf)
svmrast=sum(datastack)
svmrast[!is.na(svmrast)] <- 1
svmrast[!is.na(svmrast)] <- svm.pred
plot(svmrast)
writeRaster(svmrast, "svmrast.tif")
```

#### Lab report
This lab report is due next week (October 12th).

Produce two machine learning models using a) random forest b) another algorithm and use them to predict your response variable in your study area.

To upload in canvas:
1. A pdf with two maps produced in QGIS or ArcGIS with the predicted value for the response. Make sure to  assign an appropriate palette that enables proper visualization.

2. Figures representing variable importance for both algorithms.

3. Answer to the questions below based on your results:

a. What are the main differences between both predictions in terms of the spatial distribution of the predicted response?

b. What criteria did you use to select the final variables? Which ones did you select?

c. Are the most important variables estimated by both algorithm similar? What are the main differences? 

d. Based on the results of the partial dependence plots, what kinds of hypotheses can you produce about the correlation between the three most important variables and the response?

e. Based on your preliminary evaluation, what do you think is a better algorithm? Please justify.