---
title: 'Applied Machine Learning for Spatial Analysis. Lab Week 10: Model tunning and validation (supervised regression)'
author: "Victor Gutierrez (victorhugo@temple.edu)"
output:
  html_document:
    df_print: paged
  pdf_document: default
  word_document: default
---

```{r include = FALSE}
# Load screenshots
wd="/Users/tug61163/Library/CloudStorage/OneDrive-TempleUniversity/Repositories/Gitrepo/MALESA_Code"
setwd(wd)
library(knitr)    # For knitting document and include_graphics function
library(ggplot2)  # For plotting
library(png)      # For grabbing the dimensions of png files

fig1path=paste(wd, "Fig1.png", sep="/")
#fig2path=paste(wd, "Sel2Edit.png", sep="/")
#fig3path=paste(wd, "Triangle.png", sep="/")
#fig4path=paste(wd, "Fig4.png", sep="/")
#fig1 = readPNG(fig1path)
#fig2 = readPNG(fig2path)
#fig3 = readPNG(fig3path)
#fig4 = readPNG(fig4path)
```

## Goals
To learn how to tune and validate a supervised machine learning model.

## Lab instructions
1.	Launch R Studio and open a new R script: File/ New File/ R Script. Then save it as a new file: File/ Save As…
2.	Read the instructions below step by step. Copy and paste each chunk of code at a time in your R script. Select the code with your mouse or shift/arrow keys and then run it by pressing the keys control-enter simultaneously.

## Lab overview
For this lab, we are going to use data from previous labs to assess the most appropriate values for model hyperparameters.

The code described below was adapted from the following sources:
https://rpubs.com/phamdinhkhanh/389752
https://machinelearningmastery.com/tune-machine-learning-algorithms-in-r/
https://www.projectpro.io/recipes/tune-hyper-parameters-grid-search-r 

Have fun!

### System setup
```{r include = TRUE, message=F, warning=F, eval=FALSE}
wd=("/Users/tug61163/Library/CloudStorage/OneDrive-TempleUniversity/Courses/MALESA/2024/Week12_TunningValid")
setwd(wd)
library(caret)
```

#### Load calibration files and remove irrelevant variables
```{r include = TRUE, message=F, warning=F, eval=FALSE}
load(paste0(getwd(),"/refdata.RData"))
names(refdata)
refdata=refdata[,-2]
summary(refdata)

# Let's remove observations with NAs
refdata=subset(refdata, complete.cases(refdata))
```

### Grid search using random forest

#### Split the data in calibration and validation subsets
```{r include = TRUE, message=F, warning=F, eval=FALSE}
parts = createDataPartition(refdata$agb, p = .7, list = F)
train = refdata[parts, ]
test = refdata[-parts, ]
```

#### Setup the parameters for the k-fold cross-validation that will be passed to the train() function.
```{r include = TRUE, message=F, warning=F, eval=FALSE}
train_control = trainControl(method = "repeatedcv", 
                             number = 5, 
                             repeats=3,
                             search = "grid")
```

#### Tune the model for different values of mtry
By using the set. seed() function, you guarantee that the same random values are produced each time you run the code:  https://www.statology.org/set-seed-in-r/#:~:text=The%20set.,time%20you%20run%20the%20code.s
```{r include = TRUE, message=F, warning=F, eval=FALSE}
tunegrid <- expand.grid(.mtry = (1:5))

# For regression model
rf_tune_reg <- train(agb ~ ., 
                       data = train,
                       method = 'rf',
                       metric = 'RMSE',
                       tuneGrid = tunegrid)
print(rf_tune_reg)
plot(rf_tune_reg)
```

Some algorithms come with their own set of tools for tunning parameters. RF has a tool to optimize mtry:
```{r include = TRUE, message=F, warning=F, eval=FALSE}
y=refdata[,1]
x=refdata[, 2:ncol(refdata)]
bestMtry <- tuneRF(x,y, stepFactor = 1.5, improve = 1e-5, ntree = 500)
```

Notice that the optimum mtry is different than in the previous procedure. That might be because here the value is estimated using the OOB error estimate.

##### Create your own model by iterating hyperparameter values of ntree
```{r include = TRUE, message=F, warning=F, eval=FALSE}
tunegrid <- expand.grid(.mtry = round(c(sqrt(ncol(refdata)))))
modellist <- list()
for (ntree in c(10,100,500)){
  set.seed(123)
  fit <- train(agb~.,
               data = train,
               method = 'rf',
               metric = 'RMSE',
               tuneGrid = tunegrid,
               trControl = train_control,
               ntree = ntree)
  key <- toString(ntree)
  modellist[[key]] <- fit
}

#Compare results
results <- resamples(modellist)
summary(results)
dotplot(results)
```

#### Creating a new method to enable the tunnning of different hyperparameters. 
One option is to define a customized algorithm with a number of elements compatible with the caret package. Below is a customized algorithm to search for different combinations of mtry and ntree. This implementation can be tricky and requires experimentation to set it up. 

Here is an example with random forest. First customize the list that defines the algorithm:
```{r include = TRUE, message=F, warning=F, eval=FALSE}
metric="RMSE" ### Accuracy for supervised classification
customRF <- list(type = "Regression",
                 library = "randomForest",
                 loop = NULL)

customRF$parameters <- data.frame(parameter = c("mtry", "ntree"),
                                  class = rep("numeric", 2),
                                  label = c("mtry", "ntree"))

customRF$grid <- function(x, y, len = NULL, search = "grid") {}

customRF$fit <- function(x, y, wts, param, lev, last, weights, classProbs) {
  randomForest(x, y,
               mtry = param$mtry,
               ntree=param$ntree)
}

#Predict label
customRF$predict <- function(modelFit, newdata, preProc = NULL, submodels = NULL)
  predict(modelFit, newdata)

#Predict prob
customRF$prob <- function(modelFit, newdata, preProc = NULL, submodels = NULL)
  predict(modelFit, newdata, type = "prob")

customRF$sort <- function(x) x[order(x[,1]),]
customRF$levels <- function(x) x$classes
```

Use the custom method to train an algorithm for different values of mtry and ntree
```{r include = TRUE, message=F, warning=F, eval=FALSE}
tunegrid <- expand.grid(.mtry=c(1:15),.ntree=c(500,1000,1500))
#tunegrid <- expand.grid(.mtry=c(1:5),.ntree=c(500,1000))

set.seed(123)
custom <- train(agb~., data=train, 
                method=customRF, 
                metric=metric, 
                tuneGrid=tunegrid, 
                trControl=train_control)
print(custom)
summary(custom)
plot(custom)
```

#### Test the performance of the model to predict validation data
```{r include = TRUE, message=F, warning=F, eval=FALSE}
#use model to make predictions on test data
pred_y = predict(custom, test)

# Produce performance metrics
test_y = test[, 1]
sum(sqrt((test_y - pred_y)^2))/length(test_y) #mae - Mean Absolute Error
caret::RMSE(test_y, pred_y) #rmse - Root Mean Squared Error
```

### Grid search for xgboost
```{r include = TRUE, message=F, warning=F, eval=FALSE}
# Customizing the tuning grid
gbmGrid <-  expand.grid(max_depth = c(3, 5, 7), 
                        nrounds = (1:10)*50,    # number of trees
                        # default values below
                        eta = 0.3,
                        gamma = 0,
                        subsample = 1,
                        min_child_weight = 1,
                        colsample_bytree = 0.6)

# training a XGboost Regression tree model while tuning parameters
model = train(agb~., 
              data = train, 
              method = "xgbTree", 
              trControl = train_control, 
              tuneGrid = gbmGrid)

# summarizing the results
print(model)
plot(model)
```

##### Test the performance of the model to predict validation data
```{r include = TRUE, message=F, warning=F, eval=FALSE}
#use model to make predictions on test data
pred_y = predict(model, test)

# Produce performance metrics
test_y = test[, 1]
sum(sqrt((test_y - pred_y)^2))/length(test_y) #mae - Mean Absolute Error
caret::RMSE(test_y, pred_y) #rmse - Root Mean Squared Error
```

#### Lab report
Adapt the code illustrated today to identify suitable values for hyperparameters associated with the application of two models for the data in your final project. The first one should ba a random forest and the other an extreme gradient boosting.

Based on your results, report the optimal parameters for both algorithms based on a validation analysis and add a figure that supports your selection.

Perform an accuracy assessment for both algorithms using the test set and report your results. Based on your results, explain what  algorithm produces the best performance.