---
title: 'Applied Machine Learning for Spatial Analysis. Lab Week 7: Unsupervised classification'
author: "Victor Gutierrez (victorhugo@temple.edu)"
output:
  html_document:
    df_print: paged
  pdf_document: default
  word_document: default
---

```{r include = FALSE}
# Load screenshots
wd="/Users/tug61163/Library/CloudStorage/OneDrive-TempleUniversity/Repositories/Gitrepo/MALESA_Code"
setwd(wd)
library(knitr)    # For knitting document and include_graphics function
library(ggplot2)  # For plotting
library(png)      # For grabbing the dimensions of png files

fig1path=paste(wd, "Fig1.png", sep="/")
#fig2path=paste(wd, "Sel2Edit.png", sep="/")
#fig3path=paste(wd, "Triangle.png", sep="/")
#fig4path=paste(wd, "Fig4.png", sep="/")
#fig1 = readPNG(fig1path)
#fig2 = readPNG(fig2path)
#fig3 = readPNG(fig3path)
#fig4 = readPNG(fig4path)
```
## Lab due
October 15 2024

## Goal
To learn how to perform a spatial unsupervised classification and how to interpret the results.

## Total score
The lab counts for up to 4 points towards the final grade of the course.

## Lab instructions
1.	Launch R Studio and open a new R script: File/ New File/ R Script. Then save it as a new file: File/ Save As…
2.	Read the instructions below step by step. Copy and paste each chunk of code at a time in your R script. Select the code with your mouse or shift/arrow keys and then run it by pressing the keys control-enter simultaneously.

## Lab overview
For this lab we will implement a workflow to perform unsupervised classification using the k-means method.

The input dataset corresponds to a raster stack with two bands representing water dynamics in a wetland complex in the Colombian Caribean region between 1999 and 2019. The first raster band represents the mean proportion of the year in which a given pixel was covered by water in the time period. The second band represents the standard deviation. Data were processed from the global surface water dynamics product available here: https://glad.umd.edu/dataset/global-surface-water-dynamics

Here is the publication: https://www.sciencedirect.com/science/article/pii/S0034425720301620 

Have fun!

#### System setup
Define working directory and load required libraries. Install libraries as needed.
```{r include = TRUE, message=F, warning=F, eval=FALSE}
wd=("/Users/tug61163/Library/CloudStorage/OneDrive-TempleUniversity/Courses/MALESA/2024/Week7_Unsupervised")
setwd(wd)
dir()

library(terra)
library(sf)
library(dbscan)
```

#### Import raster dataset and plot it to make a first assessment.
```{r include = TRUE, message=F, warning=F, eval=FALSE}
waterdyn<- rast("aguavarMarialaBaja.tif")
plotRGB(waterdyn, r=1, g=2, b=1)
ext=draw("extent")
zoom(waterdyn, e=ext, maxcell=100000, layer=2, new=FALSE)
click(waterdyn, n=10, cell=TRUE)
```

#### Remove background pixels
You can see that there are some cell values equal to zero. In this case, the value has a meaning because those are areas not covered with water in any single period over the year. However, we have to remove those pixels from the analysis because they will be considered as a class in the analysis and therefore it can introduce noise and slow down the process. Let's convert any pixels with values of zero to NAs so that they are ignored from the analysis. Then plot again 
```{r include = TRUE, message=F, warning=F, eval=FALSE}
waterdyn[[2]][waterdyn[[1]]==0]=NA 
waterdyn[[1]][waterdyn[[1]]==0]=NA # I cannot remove pixels that have SD=0 because that would exclude those that have permanent water
zoom(waterdyn, e=ext, maxcell=100000, layer=1, new=FALSE)
click(waterdyn, n=10, cell=TRUE)
```

#### Convert raster into a data frame and prepare the data
Now that we successfully converted background pixels into NAs, let's convert the raster into a data frame:
```{r include = TRUE, message=F, warning=F, eval=FALSE}
df=as.data.frame(waterdyn)

# Let's do some basic exploration
hist(df[,1])
hist(df[,2])
```

K-means tends to create oblong objects and so it is not very good at handling data that comes at different scales. Therefore it is strongly adviced to re-scale all data.

A common method to rescale the data is by transforming them into zscores. From basic statstics, a zscore represents how many standard deviations a single value is away from the mean. Therefore a zscore has a mean of zero and a standard deviation of one.

We will create a function that converts values of a vector into zscores. It is a good practice to write functions at the beginning of the code. Below I add two functions, one that uses the mean and the standard deviation and the other one the median and inter-quartile range. I will use the second one because I prefer it for biased distributions but you can experiment with both and select whichever you deem the most appropriate.
```{r include = TRUE, message=F, warning=F, eval=FALSE}
zscore=function(x){
  zval=(x-mean(x, na.rm=TRUE))/sd(x, na.rm=TRUE)
  return(zval)
}

scalfun=function(x){
  xscaled=(x-median(x, na.rm=TRUE))/IQR(x, na.rm=TRUE)
  return(xscaled)
}
```

Then apply the function to all bands in the dataset.
```{r include = TRUE, message=F, warning=F, eval=FALSE}
df=apply(df, 2, scalfun) # Use zscore instead of scalfun if you want to scale based on zscores
df=data.frame(df)

summary(df)

#Let's remove any remaining entries with missing values
df=subset(df,complete.cases(df))
```

#### Run the k-means analysis using the kmeans function
```{r include = TRUE, message=F, warning=F, eval=FALSE}
clusters <- kmeans(df, centers=5, iter.max=50, nstart=5, algorithm="Lloyd")
```

The object clusters corresponds to a list storing different types of information. For now, we are concerned about the object storing the cluster number assigned to all pixels analyzed: 
```{r include = TRUE, message=F, warning=F, eval=FALSE}
head(clusters$cluster)
tail(clusters$cluster)
```

#### Rasterize results and save files into disc.
Let's create a raster with the values of pixels belonging to each cluster. For that purpose, we will first, obtain the coordinates of the analyzed pixels.
```{r include = TRUE, message=F, warning=F, eval=FALSE}
coord=terra::xyFromCell(waterdyn,(as.numeric(rownames(df))))
coord=data.frame(coord)
names(coord)=c("x", "y")

# Then create a copy of one of the layers of the original raster.
clustrast=waterdyn[[1]]
```

We will use these coordinates to identify the cell number of analyzed pixels and then replace the value of those cells in original raster for the cluster number that was assigned to them in the k-mean analysis:
```{r include = TRUE, message=F, warning=F, eval=FALSE}
xypoints=st_as_sf(coord, coords=c("x", "y"), crs=st_crs("EPSG:32618"))
xypoints=vect(xypoints)

cellval=cells(clustrast, xypoints)
clustrast[cellval[,2]]=clusters$cluster
plot(clustrast)
zoom(clustrast, e=ext, maxcell=100000, layer=1, new=FALSE)
```

Let's save the final raster and the results of the cluster analysis as a list. Let's also save the final classified raster as a tif file
```{r include = TRUE, message=F, warning=F, eval=FALSE}
clustlist=list(clustrast, clusters)
save(clustlist, file='cluster_Lloyd.RData')
writeRaster(clustrast, 'cluster_Lloyd.tif')
```

#### Interpret clustering results
Let's open both clustering results in QGIS. We will see how to overlay the results with a high resolution image downloaded from google earth. Let's retrieve the centroids of the clusters and use them to interpret the classified map
```{r include = TRUE, message=F, warning=F, eval=FALSE}
clusters$centers
```

#### Produce Density-Based Spatial Clustering. Warning: You might run out of memory!
```{r include = TRUE, message=F, warning=F, eval=FALSE}
dbClust <- dbscan(df, eps = 0.5, minPts = 100)  # You can try other methods like "complete", "single", etc.

# Plot the dendrogram
print(dbClust)

# Interpret the results
dbscan_result$cluster
```

#### Include coordinates as variables in the cluster
Let's also explore these results in QGIS
```{r include = TRUE, message=F, warning=F, eval=FALSE}
df$x=scalfun(coord[,1])
df$y=scalfun(coord[,2])

for(xval in (ncol(df)-1):ncol(df)){
    df[,xval]=scalfun(df[,xval])
  }
clustersGeo <- kmeans(df, centers=5, iter.max=50, nstart=5, algorithm="Lloyd")
clustrastGeo=waterdyn[[1]]
clustrastGeo[cellval[,2]]=clustersGeo$cluster
plot(clustrastGeo)
zoom(clustrastGeo, e=ext, maxcell=100000, layer=1, new=FALSE)
clustlistGeo=list(clustrast, clustersGeo)
save(clustlistGeo, file='clusterGeo_Lloyd.RData')
writeRaster(clustrastGeo, 'clusterGeo_Lloyd.tif')
clustersGeo$centers
```

## Lab instructions

1. Select three to four variables that you believe, will likely have the strongest influence on the response that you want to predict for your final project. Briefly describe the rationale for the selection of the variables based on the results of your EDA.

2. Run a k-means analysis, save the raster into the disc and map the results in QGIS. Interpret the mapping results based on your understanding of the phenomenon that you want to predict and the values of the centroids obtained in the k-means analysis.

3. Upload to canvas: 
a. The answers to points 1 and 2
b. A screenshot of the classified map, with unique values for each class and another screenshot for the same area displaying a high resolution image from google earth.
c.  the values of the centroids obtained with your analysis for the outputs that you selected as the best (with or without map coordinates).




