---
title: 'Applied Machine Learning for Spatial Analysis. Lab Week 10: Supervised classification'
author: "Victor Gutierrez (victorhugo@temple.edu)"
output:
  html_document:
    df_print: paged
  pdf_document: default
  word_document: default
---

```{r include = FALSE}
# Load screenshots
wd="/Users/tug61163/Library/CloudStorage/OneDrive-TempleUniversity/Repositories/Gitrepo/MALESA_Code"
setwd(wd)
library(knitr)    # For knitting document and include_graphics function
library(ggplot2)  # For plotting
library(png)      # For grabbing the dimensions of png files

fig1path=paste(wd, "Fig1.png", sep="/")
#fig2path=paste(wd, "Sel2Edit.png", sep="/")
#fig3path=paste(wd, "Triangle.png", sep="/")
#fig4path=paste(wd, "Fig4.png", sep="/")
#fig1 = readPNG(fig1path)
#fig2 = readPNG(fig2path)
#fig3 = readPNG(fig3path)
#fig4 = readPNG(fig4path)
```

## Goals
To learn:
1. How to fit and interpret a supervised classification using tree based methods and 
2. How to use the fitted model to produce spatial predictions.

## Lab instructions
1.	Launch R Studio and open a new R script: File/ New File/ R Script. Then save it as a new file: File/ Save As…
2.	Read the instructions below step by step. Copy and paste each chunk of code at a time in your R script. Select the code with your mouse or shift/arrow keys and then run it by pressing the keys control-enter simultaneously.

## Lab overview
For this lab, we are going to use three machine learning methods to predict the distribution of four vegetation communities in Kyrgyzstan. Predictors or "features" correspond to spectral data obtained from Landsat imagery, distance to settlements, and several topographical indices.


Have fun!

### 1. System setup
```{r include = TRUE, message=F, warning=F, eval=FALSE}
wd=("//Users/tug61163/Library/CloudStorage/OneDrive-TempleUniversity/Courses/MALESA/2024/Week10_SuperClass")
setwd(wd)
library(terra)
library(sf)
library(ggplot2)
library(DataExplorer)
library(randomForest)
library(xgboost)
library(rpart)
library(rattle)
library(pdp)
library(iml)
```

#### Load and prepare spatial files
```{r include = TRUE, message=F, warning=F, eval=FALSE}
# Stack of raster predictors:
kgz=terra::rast("RasterStack_KGZ.tiff")

# Scale dataset so that the magnitude of variables is comparable
kgzScaled=terra::scale(kgz)
names(kgzScaled) = c("Aspect", "B2", "B3", "B4", "B5",
               "B6", "B7", "DEM", "SettDist", 
               "LS", "NDVI", "NDWI", "SVF",
               "Slope", "TWI", "VD")

# B2-B7: ground reflectance data for bands 1 through seven obtained from Landsat data
# SettDist: distance to settlements
# LSF: LS_Factor
# SVF: Sky View Factor
# TWI: Topographic Wetness Index
# VD: Valley Depth

# Ground data representing the response variable as points
unzip("VegPlots.zip")
classes=st_read("Vegetation Plots.shp") # open in sf()
classes=terra::vect(classes)  # Transform to terra()     
```

#### Extract data to dataframe and prepare for modeling
```{r include = TRUE, message=F, warning=F, eval=FALSE}
# Extract data into a dataframe
sampdata=terra::extract(kgzScaled, classes, xy=TRUE)

# add response variable to the datafame
sampdata$veg=classes$Vegetation
sampdata=subset(sampdata, complete.cases(sampdata))
sampdata$veg=as.factor(sampdata$veg)
names(sampdata)

# make sure that the names of the stack are the same as the names of the dataframe. This is important to make sure the prediction works!
names(kgzScaled)==names(sampdata[-c(1,18:20)])

#Subset data before modeling to remove irrelevant variables and also to make the variables in the dataframe match the variables in the raster stack
#names(sampdata)==names(kgz)
#sampdata=sampdata[,-c(1, 18:19)]
```

#### Explore the data
```{r include = TRUE, message=F, warning=F, eval=FALSE}
dfSummary(refdata) # Check data structure
sampdata %>%
    create_report(
        output_file = paste("Report", format(Sys.time(), "%Y-%m-%d %H:%M:%S %Z"), sep=" - "),
        report_title = "EDA Report - Plant communities",
        y = "veg"
    )
```

# Band selection
We will do variable selection as part of the class excercise
```{r include = TRUE, message=F, warning=F, eval=FALSE}
plot_bar(sampdata)

ggplot(sampdata, aes(x = veg, y = DEM)) + geom_boxplot()
ggplot(sampdata, aes(x = veg, y = NDWI)) + geom_boxplot()
ggplot(sampdata, aes(x = veg, y = B5_Ref)) + geom_boxplot()
ggplot(sampdata, aes(x = veg, y = Slope)) + geom_boxplot()
ggplot(sampdata, aes(x = veg, y = SettDist)) + geom_boxplot()
ggplot(sampdata, aes(x = veg, y = SVF)) + geom_boxplot()
ggplot(sampdata, aes(x = veg, y = VD)) + geom_boxplot()

plot_correlation(sampdata)
```

#### Fit different ML methods and interpret results

##### Decision trees
```{r include = TRUE, message=F, warning=F, eval=FALSE}
dt=rpart(veg~DEM + NDWI + B5 + Slope + SettDist + SVF + VD, data=sampdata, method="class")
# Check the arguments for rpart.control
# Explain the tree outputs

printcp(dt) #displays the results
plotcp(dt)
summary(dt)

dt$variable.importance
plot(dt, uniform=TRUE, 
     main="Classification Tree for plants")
text(dt, use.n=TRUE, all=TRUE, cex=.8)

fancyRpartPlot(dt, caption = NULL)

printcp(dt)

#Pruning the tree
dt$cptable # complexity table. 
# The first element represents the levels in the tree
# The second element shows the complexity parameter. Lower values correspond to more complex trees
# N split corresponds to the number of nodes in the tree
# rel error is the error of the classification at diferent nodes
# x error is the expected error if the model was applied to new data
# xstd is the standard deviation of xerror
plotcp(dt)
dtprune=prune(dt, cp= dt$cptable[which.min(dt$cptable[,"xerror"]),"CP"])

# plot the pruned tree 
plot(dtprune, uniform=TRUE, 
     main="Pruned Classification Tree for vegetation communities")
text(dtprune, use.n=TRUE, all=TRUE, cex=.8)
```

##### Random forest
```{r include = TRUE, message=F, warning=F, eval=FALSE}
rf=randomForest(veg~DEM + NDWI + B5 + Slope + SettDist + SVF + VD, data=sampdata, ntree=200, importance=TRUE)
rf
rf$importance
varImpPlot(rf)
```

#####  Extreme Boosting machine 
https://xgboost.readthedocs.io/en/stable/tutorials/model.html 
```{r include = TRUE, message=F, warning=F, eval=FALSE}
sampdataX=sampdata[c("B5", "DEM", "SettDist", "NDWI", "SVF", "Slope", "VD")]
names(sampdataX)=c("B5", "DEM", "SettDist","NDWI", "SVF", "Slope", "VD")
#sampdataX_scaled <- scale(sampdataX)
veg=sampdata$veg

#I have to convert variables into an xgb matrix. xgb only accepts integers starting from zero for multiclass models
labels = as.numeric(veg)-1 # The names of the vegetation types are converted into numbers starting from zero in alphabetical order

dtrain <- xgb.DMatrix(data = as.matrix(sampdataX), label = labels, missing = NA)

# Set parameters for the model
param <- list(
  objective = "multi:softprob",  # Multiclass classification
  num_class = 4,  # Adjust based on the number of classes
  eta = 0.1,  # learning rate
  max_depth = 6, # maximum depth of the tree
  eval_metric = "merror"  # Error rate
)

# Train the model using xgb.train()
xgb_model <- xgb.train(params = param, data = dtrain, nrounds = 500, nthread = 6)

# Variable importance
importance <- xgb.importance(feature_names = colnames(sampdataX), model = xgb_model)
head(importance)
xgb.plot.importance(importance_matrix = importance)
```

#### Evaluate marginal effects

##### For random forest
```{r include = TRUE, message=F, warning=F, eval=FALSE}
### For one predictor

# Define predictor object for the model
predictor_rf <- Predictor$new(rf, data = sampdataX, y = veg)

# Plot marginal effects for a specific feature
effect <- FeatureEffect$new(predictor_rf, feature = "Slope", method = "pdp")
plot(effect)

effect2 <- FeatureEffect$new(predictor_rf, feature = "DEM", method = "pdp")
plot(effect2)

effect3 <- FeatureEffect$new(predictor_rf, feature = "B5", method = "pdp")
plot(effect3)

### For two predictors
pdp_two_features_rf <- FeatureEffect$new(predictor_rf, 
                                         feature = c("B5", "NDWI"), 
                                         method = "pdp")

# Plot the 2D PDP
plot(pdp_two_features_rf)
```

##### For xgboost
```{r include = TRUE, message=F, warning=F, eval=FALSE}
# Prepare the Predictor object from iml package
predictor_xgb <- Predictor$new(xgb_model, data = as.data.frame(sampdataX), y = labels, 
                               predict.function = function(model, newdata) {
                                 matrix(predict(model, as.matrix(newdata)), ncol = 4, byrow = TRUE)
                               })

# Create a partial dependence plot for a single feature
pdp_iml <- FeatureEffect$new(predictor_xgb, feature = "B5", method = "pdp")
plot(pdp_iml)

### For two predictors
xgb_pred <- function(model, newdata) {
  preds <- predict(model, as.matrix(newdata))
  preds <- matrix(preds, ncol = 4, byrow = TRUE)  # Reshape the predictions
  return(preds)  # Returns the predicted probabilities for each class
}

predictor_xgb <- Predictor$new(xgb_model, 
                               data = as.data.frame(train_data), 
                               y = train_labels, 
                               predict.function = xgb_pred)

# Generate a PDP for two predictors
pdp_two_features <- FeatureEffect$new(predictor_xgb, 
                                      feature = c("B5", "NDWI"), 
                                      method = "pdp")

# Plot the 2D PDP
plot(pdp_two_features)
```

#### Produce predicted maps and save them
```{r include = TRUE, message=F, warning=F, eval=FALSE}
### For decision trees

dtpred=terra::predict(kgzScaled, dtprune)
plot(dtpred)

# Let's mask out invalid pixels
msk=sum(kgz)
msk[!is.na(msk)]<-1
plot(msk)

dtpred=dtpred*msk
plot(dtpred)

# To produce a single-layer raster where each pixel represents the layer with the highest probability:
# Extract the values from the raster stack as a matrix
dtpredVal <- values(dtpred)

# Determine the index of the layer with the highest probability for each pixel
# max.col finds the column index (i.e., layer) with the max value for each row (i.e., pixel)
dtpredmaxVal <- max.col(dtpredVal, ties.method = "first")

# Create a new raster to store the result
dtpredRast <-msk  # Use one layer to initialize the raster

# Assign the layer indices to the new raster
values(dtpredRast) <- dtpredmaxVal
plot(dtpredRast)


### For Random Forest
rfpred=terra::predict(kgzScaled, rf)
plot(rfpred)

### For xgb
# Subset kgz with only the variables used to calibrate the model
kgzSub=kgzScaled[[c("B5", "DEM", "SettDist", "NDWI", "SVF", "Slope", "VD")]]
plot(kgzSub)

# Let's mask (kgzSub) to remove invalid pixels
kgzSub=kgzSub*msk
plot(kgzSub)

# Convert kgz to matrix
kgz_matrix <- as.matrix(kgzSub)
length(which(!is.na(kgz_matrix)==TRUE))
kgz_matrix=subset(kgz_matrix, complete.cases(kgz_matrix))

#xgbpred <- predict(xgb_model, as.matrix(kgzSub))
xgbpred <- predict(xgb_model, kgz_matrix)

# Reshape the prediction output to have one row per observation and one column per class
xgbpred_matrix <- matrix(xgbpred, nrow = nrow(kgz_matrix), ncol = 4, byrow = TRUE)

# For each observation, select the class with the highest probability
predicted_classes <- max.col(xgbpred_matrix)  # Subtract 1 because classes are 0-indexed

# Make predictions

# Create a raster template
xgRast=sum(kgzSub)

# make sure that the number of valid pixels in the template is equal to the number of predicted values
length(xgRast[!is.na(xgRast)])== length(predicted_classes)

# Replace value for valid pixels to the predicted values
xgRast[!is.na(xgRast)] <- predicted_classes
# Count how many pixels are not NA
plot(xgRast)

writeRaster(dtpred, "dtpred.tif")
writeRaster(rfpred, "rfpred.tif")
writeRaster(xgRast, "xgbRast.tif")
```

#### Lab report
This lab report is due in two weeks (October 12th), after covering supervised regression.

Produce two machine learning models using a) random forest b) another algorithm and use them to predict your response variable in your study area.

To upload in canvas:
1. A pdf with two maps produced in QGIS or ArcGIS with the predicted value for the response. Make sure to  assign an appropriate palette that enables proper visualization.

2. Figures representing variable importance for both algorithms.

3. Answer to the questions below based on your results:

a. What are the main differences between both predictions in terms of the spatial distribution of the predicted response?

b. What criteria did you use to select the final variables? Which ones did you select?

c. Are the most important variables estimated by both algorithm similar? What are the main differences? 

d. Based on the results of the partial dependence plots, what kinds of hypotheses can you produce about the correlation between the three most important variables and the response?

e. Based on your preliminary evaluation, what do you think is a better algorithm? Please justify.