---
title: 'Applied Machine Learning for Spatial Analysis. Lab Week 11: Model tunning and validation (supervised regression)'
author: "Victor Gutierrez (victorhugo@temple.edu)"
output:
  html_document:
    df_print: paged
  pdf_document: default
  word_document: default
---

```{r include = FALSE}
# Load screenshots
wd="/Users/tug61163/Library/CloudStorage/OneDrive-TempleUniversity/Repositories/Gitrepo/MALESA_Code"
setwd(wd)
library(knitr)    # For knitting document and include_graphics function
library(ggplot2)  # For plotting
library(png)      # For grabbing the dimensions of png files

fig1path=paste(wd, "Fig1.png", sep="/")
#fig2path=paste(wd, "Sel2Edit.png", sep="/")
#fig3path=paste(wd, "Triangle.png", sep="/")
#fig4path=paste(wd, "Fig4.png", sep="/")
#fig1 = readPNG(fig1path)
#fig2 = readPNG(fig2path)
#fig3 = readPNG(fig3path)
#fig4 = readPNG(fig4path)
```

## Goals
To learn how to tune and validate a supervised machine learning model.

## Lab instructions
1.	Launch R Studio and open a new R script: File/ New File/ R Script. Then save it as a new file: File/ Save As…
2.	Read the instructions below step by step. Copy and paste each chunk of code at a time in your R script. Select the code with your mouse or shift/arrow keys and then run it by pressing the keys control-enter simultaneously.

## Lab overview
For this lab, we are going to use data from previous labs to assess the most appropriate values for model hyperparameters and then apply them to test data for model selection

Have fun!

### 1.System setup, loading required libraries and files
```{r include = TRUE, message=F, warning=F, eval=FALSE}
wd=("/Users/tug61163/Library/CloudStorage/OneDrive-TempleUniversity/Courses/MALESA/2024/Week12_TunningValid")
setwd(wd)
library(tidymodels)
library(tune)
library(dials)
library(workflows)
library(parsnip)
library(rsample)
library(recipes)
library(yardstick)
library(ranger)

datastack=terra::rast("RasterStack_KGZ.tiff")
datastackScaled=terra::scale(kgz)
names(datastackScaled) = c("Aspect", "B2", "B3", "B4", "B5",
               "B6", "B7", "DEM", "SettDist", 
               "LS", "NDVI", "NDWI", "SVF",
               "Slope", "TWI", "VD")

load(paste0(getwd(),"/sampdata.RData"))
summary(sampdata)
names(sampdata)[2:17] = c("Aspect", "B2", "B3", "B4", "B5",
                     "B6", "B7", "DEM", "SettDist", 
                     "LS", "NDVI", "NDWI", "SVF",
                     "Slope", "TWI", "VD")
        
sampdata=sampdata[,c(2:17,20)] # remove the line that corresponds to ID, x, y
summary(sampdata)
```

### 2.  Split for Training, Validation, and Testing
```{r include = TRUE, message=F, warning=F, eval=FALSE}
set.seed(123)
data_split <- initial_split(sampdata, prop = 0.8)
train_data_final <- training(data_split)
test_data <- testing(data_split)
```

### 3. Define a Recipe for Preprocessing
```{r include = TRUE, message=F, warning=F, eval=FALSE}
recipe_veg <- recipe(veg ~ ., data = train_data_final) %>%
  # step_normalize(all_predictors()) %>%
   step_naomit(all_predictors())
```

### 4. Define Model Specifications
```{r include = TRUE, message=F, warning=F, eval=FALSE}
# 4.1 Random Forest Model
# Define the Random Forest model specification with tuning parameters
rf_spec <- rand_forest(
  mtry = tune(),        # Number of predictors randomly sampled at each split
  min_n = tune(),       # Minimum number of data points in a node to make a split
  trees = tune()        # Number of trees in the forest
) %>%
  set_mode("classification") %>%
  #set_engine("ranger")
  set_engine("randomForest")

# An engine is basically a package that represents an implementation of a machine learning method.
#You can see what are the engines available for random forest:
show_engines("rand_forest")

# We select "ranger" for its efficiency and capacity to handle large datasets

# 4.2 Extreme Gradient Boosting Model
xgb_spec <- boost_tree(
  trees = tune(),          # Number of trees
  tree_depth = tune(),     # Maximum depth of each tree
  learn_rate = tune(),     # Learning rate
  loss_reduction = tune(), # Minimum reduction in loss for a split
  sample_size = tune(),    # Fraction of observations to sample
  mtry = tune()            # Number of predictors to randomly sample at each split
) %>%
  set_mode("classification") %>%
  set_engine("xgboost") 

# 4.3 Neural Network Model
nn_spec <- mlp(hidden_units = tune(), penalty = tune(), epochs = tune()) %>%
  set_mode("classification") %>%
  set_engine("nnet")
```

### 5. Create Workflows
```{r include = TRUE, message=F, warning=F, eval=FALSE}
# Create workflows for each model
rf_workflow <- workflow() %>%
  add_model(rf_spec) %>%
  add_recipe(recipe_veg)

xgb_workflow <- workflow() %>%
  add_model(xgb_spec) %>%
  add_recipe(recipe_veg)

nn_workflow <- workflow() %>%
  add_model(nn_spec) %>%
  add_recipe(recipe_veg)
```

### 6. Set Up K-Fold Cross-Validation
```{r include = TRUE, message=F, warning=F, eval=FALSE}
set.seed(123)
cv_folds <- vfold_cv(train_data_final, v = 5)
```

### 7. Define Tuning Grids for Each Model
```{r include = TRUE, message=F, warning=F, eval=FALSE}
rf_grid <- grid_random(finalize(mtry(), train_data_final), min_n(), trees(), size = 20)

# Define the tuning grid for Random Forest
rf_grid <- grid_random(
  mtry(range = c(2, 10)),      # Limits mtry between 2 and 10
  min_n(range = c(5, 15)),     # Limits min_n between 5 and 15
  trees(range = c(100, 1000)), # Limits trees between 100 and 1000
  size = 20
)

# You can instead use the function to test all possible combinations of paraeters but this would take way longer
# rf_grid <- grid_regular(
#   mtry(range = c(2, 10)),  # Divides mtry into a specified number of levels
#   trees(range = c(50, 500)),
#   min_n(range = c(2, 15)),
#   levels = 5               # Divides each parameter range into 5 levels
# )

xgb_grid <- grid_random(
  finalize(mtry(), train_data_final), 
  trees(),
  tree_depth(),
  learn_rate(),
  loss_reduction(),
  sample_prop(),  # Use sample_size() to match xgb_spec
  size = 20
)

nn_grid <- grid_random(hidden_units(), penalty(), epochs(), size = 20)
```

### 8. Tune Models Using Cross-Validation
```{r include = TRUE, message=F, warning=F, eval=FALSE}
# 8.1 Tune Random Forest
set.seed(123)
rf_tuned <- tune_grid(
  rf_workflow,
  resamples = cv_folds,
  grid = rf_grid,
  metrics = metric_set(bal_accuracy)
)

# 8.2 Tune Extreme Gradient Boosting
set.seed(123)
xgb_tuned <- tune_grid(
  xgb_workflow,
  resamples = cv_folds,
  grid = xgb_grid,
  metrics = metric_set(bal_accuracy)
)

#8.3 Tune Neural Network
set.seed(123)
nn_tuned <- tune_grid(
  nn_workflow,
  resamples = cv_folds,
  grid = nn_grid,
  metrics = metric_set(bal_accuracy)
)
```


### 9. Select the Best Hyperparameters Based on balanced accuracy
```{r include = TRUE, message=F, warning=F, eval=FALSE}
show_best(rf_tuned, metric="bal_accuracy") # this shows the top 5 models
rf_best <- select_best(rf_tuned, metric="bal_accuracy") # this selects the very best model
rf_best_simp <- select_by_one_std_err(rf_tuned, metric="bal_accuracy", min_n)

show_best(xgb_tuned, metric="bal_accuracy") 
xgb_best <- select_best(xgb_tuned, metric="bal_accuracy")

show_best(nn_tuned, metric="bal_accuracy")
nn_best <- select_best(nn_tuned, metric="bal_accuracy")
```

### 10. Finalize the Workflows with Optimal Hyperparameters
```{r include = TRUE, message=F, warning=F, eval=FALSE}
rf_workflow <- finalize_workflow(rf_workflow, rf_best)
xgb_workflow <- finalize_workflow(xgb_workflow, xgb_best)
nn_workflow <- finalize_workflow(nn_workflow, nn_best)
```

# Step 11: Fit Final Models on Training Data and Evaluate on Test Data
```{r include = TRUE, message=F, warning=F, eval=FALSE}
# Random Forest
rf_final <- rf_workflow %>%
  last_fit(data_split)
rf_metrics <- rf_final %>%
  collect_metrics()

# Extreme Gradient Boosting
xgb_final <- xgb_workflow %>%
  last_fit(data_split)
xgb_metrics <- xgb_final %>%
  collect_metrics()

# Neural Network
nn_final <- nn_workflow %>%
  last_fit(data_split)
nn_metrics <- nn_final %>%
  collect_metrics()
```

### 12. Display metrics for each model, compare Model Performance and select best model
```{r include = TRUE, message=F, warning=F, eval=FALSE}
rf_metrics
xgb_metrics
nn_metrics
```

#### Lab report
Adapt the code illustrated today to identify suitable values for hyperparameters associated with the application of the three models for the data in your final project. 

Perform an accuracy assessment for both algorithms using the test set and report your results. Based on your results, explain what algorithm produces the best performance.