---
title: 'Applied Machine Learning for Spatial Analysis. Lab Week 2: Introduction to R Environment and Structures'
author: "Victor Gutierrez (victorhugo@temple.edu)"
output:
  html_document:
    df_print: paged
  pdf_document: default
  word_document: default
---

```{r include = FALSE}
# Load screenshots
wd="/Users/tug61163/Library/CloudStorage/OneDrive-TempleUniversity/Repositories/Gitrepo/MALESA_Code"
library(knitr)    # For knitting document and include_graphics function
library(ggplot2)  # For plotting
library(png)      # For grabbing the dimensions of png files
setwd(wd)
fig1path=paste(wd, "Fig1.png", sep="/")
#fig2path=paste(wd, "Sel2Edit.png", sep="/")
#fig3path=paste(wd, "Triangle.png", sep="/")
#fig4path=paste(wd, "Fig4.png", sep="/")
#fig1 = readPNG(fig1path)
#fig2 = readPNG(fig2path)
#fig3 = readPNG(fig3path)
#fig4 = readPNG(fig4path)
```
## Lab due
There is no lab report due for this session.

## Goal
To become familiar with the R environment as well as basic data structures.

## Total score
This lab is not scored

### Downloading and opening required software
The installation instructions below are adapted from here (https://www.andrewheiss.com/blog/2012/04/17/install-r-rstudio-r-commander-windows-osx/)

### In Windows
1.	Download R from http://cran.us.r-project.org/ (click on “Download R for Windows” > “base” > “Download R X.X.X for Windows” (latest version). You can also click on "Installation and other instructions for further details.
2. Follow the instructions to install R. Leave all default settings in the installation options.
3.	Download RStudio from http://rstudio.org/download/desktop and install it. Leave all default settings in the installation options.

### In Mac
1.	Download R from http://cran.us.r-project.org/ (click on “Download R for Mac OS X” > “R-X.X.X.pkg” (latest version).
2.	Follow the instructions to install R. Leave all default settings in the installation options.
3.	Download RStudio from http://rstudio.org/download/desktop.
4.	Install RStudio by double clicking on the downloaded file.
5. Install xcode to execute the plot3D function. You can download it from here: https://developer.apple.com/xcode/ . You will need to restart your machine after installation.
6. Install homebrew. Open terminal and paste the line of code that appears on top here: https://brew.sh
7. Install the cartographic projections library by pasting and running the line of code that appears in the following link in your terminal: https://formulae.brew.sh/formula/proj

## Lab instructions
1.	Launch R Studio and open a new R script: File/ New File/ R Script. Then save it as a new file: File/ Save As…
2.	Read the instructions below step by step. Copy and paste each chunk of code at a time in your R script. Select the code with your mouse or shift/arrow keys and then run it by pressing the keys control-enter simultaneously.

I strongly recommend this introductory R book: https://cran.r-project.org/doc/manuals/r-release/R-intro.pdf 

Have fun!

### 1. The RStudio Environment
R studio has four main panels.
1. Code: This is where you write and save your scripts
2. Console: Here is where you run the code and can visualize the outputs of operations.
3. Environment: You can visualize here the objects loaded in R
4. Files, plots, packages, help: Use it to explore folders, visualize plots, load libraries and document the execution of functions.

### 2. R essentials
You can write code directly in the console or run it using the code panel. To run code using the code panel, you can locate the cursor in the line of code that you want to run or select a chunk of code (shift+arrows) and then execute it by typing shift+enter
```{r include = TRUE, message=F, warning=F, eval=FALSE}
"hello world"
```

Or make basic operations
```{r include = TRUE, message=F, warning=F, eval=FALSE}
2+3
```

R operates with objects. For instance, instead of writing 2+3, you can do:
```{r include = TRUE, message=F, warning=F, eval=FALSE}
a=2
b=3
c=a+b
a
b
c
```

In this case, a, b, and c are objects (notice that they are loaded in the environment panel). You can re-use objects for further operations
```{r include = TRUE, message=F, warning=F, eval=FALSE}
c^(a/b)
```

Objects can be numeric or strings (text)
```{r include = TRUE, message=F, warning=F, eval=FALSE}
d = "I love"
e = "machine learning"
```

In addition to numeric or character there are also logical values in R. You can know the types of values or vectors by using the function class()
```{r include = TRUE, message=F, warning=F, eval=FALSE}
class(a)
class(b)
a>b
class(a>b)
```

There are other types of values that are considered as missing values or not a number (NaN). Any operation with these values return an (NaN)
```{r include = TRUE, message=F, warning=F, eval=FALSE}
h=sqrt(-c)
h
a*h
```

There are also infinity numbers. Any operated with infinity results into infinity, except for a NaN
```{r include = TRUE, message=F, warning=F, eval=FALSE}
i=1/0
a*i
h*i
```

### 3. Data structures

#### Vectors
You can  store different elements in a data structure known as "vector". A vector is created using the function c
```{r include = TRUE, message=F, warning=F, eval=FALSE}
vec1=c(10, 5, 3, 9, 4)
vec1
vec2=c(a, b, c, a/b, b*c)
vec2
vec3=c(vec1, vec2)
vec3
```

In fact a, b, and c are considered by R as a vector with only one element. You can know the number of elements in a vector by using the function length
```{r include = TRUE, message=F, warning=F, eval=FALSE}
length(a)
length(vec1)
```

An operation between a number and a vector, returns an object that applies the value of the number to all the elements in the vector. 
```{r include = TRUE, message=F, warning=F, eval=FALSE}
a*vec1
```

An operation between vectors of different lengths results in repeating some of the elements in the shorter vector until the length matches the one of the longest vector
```{r include = TRUE, message=F, warning=F, eval=FALSE}
vec4=c(5,2,9)
vec3+vec4
```

Vectors are indexed so you can select given elements from a vector for different purposes
```{r include = TRUE, message=F, warning=F, eval=FALSE}
vec1[c(1,3,5)]
vec1>4
vec1[vec1>4]
vec1[vec1>4]=NA
```

Vectors are considered as a class on their own. Other classes in R are matrix, array, factor, data frame and list

#### Matrices
Matrices are objects that have two dimensions
```{r include = TRUE, message=F, warning=F, eval=FALSE}
m=matrix(seq(1,9), nrow=3) # is the same as
m=matrix(seq(1,9), ncol=3)
m
```

Matrices are usually arranged by column but you can also arrange them by row with an additional argument in the matrix() function
```{r include = TRUE, message=F, warning=F, eval=FALSE}
m=matrix(seq(1,9), ncol=3, byrow=T)
m
```

Matrices can be indexed by using the row and column number
```{r include = TRUE, message=F, warning=F, eval=FALSE}
m[2,2]
m[2,]
m[,2]
class(m)
```

You can also join two vectors to produce a matrix
```{r include = TRUE, message=F, warning=F, eval=FALSE}
m2=cbind(vec1,vec2)
m3=rbind(vec1,vec2)
class(m2)
class(m3)
```

#### Arrays
Arrays are objects with one or more dimensions. A matrix is an array with only two dimensions
```{r include = TRUE, message=F, warning=F, eval=FALSE}
ar=array(seq(1,36), c(3,4,3))
ar
ar[2,3,]
m4=ar[1:3, 1:2, 3]
class(ar)
class(m4)
```

#### Factors
A factor is a data structure used to store categorical variables. Some machine learning algorithms require some variables to be entered as factors so that they recognize entries with the same name as part of the same class.
```{r include = TRUE, message=F, warning=F, eval=FALSE}
LandCover=c("grassland","impervious", "forest", "forest", "water", "grassland", "bare soil", "impervious", "grassland")
LandCover
class(LandCover)
LandCover=factor(LandCover)
LandCover
class(LandCover)
```

#### Data frames
Data frames are the most used type of structures for machine learning in R. They are typically used to store observations (arranged as rows) for different types of variables (arranged as columns). They are the equivalent of a spreadsheet in R. They can also be understood as a set of vectors or factors with the same length
```{r include = TRUE, message=F, warning=F, eval=FALSE}
temperature=c(25, 32, 18, 21, 12, 27, 29, 33, 26)
humidity=c(17, 6, 55, 67, 100, 22, 12, 8, 19)
soildepth=c(20, 0, 70, 65, 0, 17, 12, 0, 19)
ecodata=data.frame(LandCover, temperature, humidity, soildepth)
class(ecodata)
```

You can actually load a data frame in R studio as a spreadsheet by double clicking on its name in the environment panel.

#### Lists
A list is the most flexible type of object in R. It can compile different types of data sets in different compartments or slots. Due to this flexibility, lists are often used to store various types of input and output data and sets of configuration parameters for machine learning models.
```{r include = TRUE, message=F, warning=F, eval=FALSE}
l=list(a, paste(d,e, sep=" "), h, vec1, ar, ecodata)
l
class(l)
```

You can retrieve an element of the list by using double bracekt
```{r include = TRUE, message=F, warning=F, eval=FALSE}
l[[6]]
l[[6]][,3]
```

#### Functions
Functions are pre-coded computations applied to objects. For instance:
```{r include = TRUE, message=F, warning=F, eval=FALSE}
sqrt(a)
f=sum(vec3)
g=length(vec3)
mean(vec3)
f/g
sd(vec3)
sort(vec1)
paste(d,e, sep=" ")
```

You can see the code behind each function by writing the name of the function:
```{r include = TRUE, message=F, warning=F, eval=FALSE}
sqrt
sd
paste
```

Functions are implemented through the use of arguments. Each function has its own arguments. Some of them are required and some of them are optional. You can learn how to implement a function by writing ? followed by the name of the function
```{r include = TRUE, message=F, warning=F, eval=FALSE}
?paste
```

Use the function rm() to remove objects from the environment. Check that removed objects disappear from the "Environment" panel
```{r include = TRUE, message=F, warning=F, eval=FALSE}
rm(l)
rm(list=ls()) # removes all objects from the environment
```



