---
title: 'Advanced Remote Sensing Lab 3: Spatial segmentation, object-based analyses, and contextual metrics'
author: "Victor Gutierrez (victorhugo@temple.edu)"
output:
  html_document:
    df_print: paged
  word_document: default
  pdf_document: default
---

### Lab due
Febraury 16 2021

### Goals
1. 

### Total score
The lab counts for up to 3 points towards the final grade of the course.

Have fun!

```{r include = FALSE}
# Load screenshots
wd="//Users/tug61163/Documents/Courses/AdvancedRS/Spring2021/Session4_SegmentationTextural"
library(knitr)    # For knitting document and include_graphics function
library(ggplot2)  # For plotting
library(png)      # For grabbing the dimensions of png files
setwd(wd)
fig1path=paste(wd, "SelEdit.png", sep="/")
fig2path=paste(wd, "Sel2Edit.png", sep="/")
fig3path=paste(wd, "Triangle.png", sep="/")
#fig4path=paste(wd, "Fig4.png", sep="/")
#fig1 = readPNG(fig1path)
#fig2 = readPNG(fig2path)
#fig3 = readPNG(fig3path)
#fig4 = readPNG(fig4path)
```

### Lab instructions

### 1. Environment setup and data loading
Use the image corresponding to the spectral unmixing results from lab 3. You can download mine from the session 4 module in canvas: PENDING

Save downloaded image in a folder that you will use as a working directory. 

Setup working directory and load required libraries. In my case I saved it here: "/Users/tug61163/Documents/Courses/AdvancedRS/Spring2021/Session4_SegmentationTextural".
```{r include = TRUE, message=F, warning=F, eval=FALSE}
setwd("/Users/tug61163/Documents/Courses/AdvancedRS/Spring2021/Session4_SegmentationTextural")
library(raster)
library(OpenImageR)
library(randomForest)
library(RStoolbox)
library(dplyr)
```

Open and plot image.
```{r include = TRUE, message=F, warning=F, eval=FALSE}
l8unmix=stack("unmixPCA.tif")
plotRGB(l8unmix, r=1, g=2, b=3, stretch="lin")
```

#### 2. Segment the image
The package to be used for segmentation only recognizes images in JPEG format. Here we will convert the image from raster to JPEG format while conserving the original dimensionality.
```{r include = TRUE, message=F, warning=F, eval=FALSE}
nrows=l8unmix@nrows
ncols=l8unmix@ncols
jpeg("Unmix.jpg", width = ncols, height = nrows)
plotRGB(l8unmix, r=1, g=2, b=3, stretch="lin")
dev.off()
```

We will use the super-pixels method. Modify the superpixel argument until you obtain satisfactory results. Larger values translate into more and smaller superpixels.
```{r include = TRUE, message=F, warning=F, eval=FALSE}
rgb <- readImage("Unmix.jpg")
seg = superpixels(input_image =rgb, method = "slic", superpixel = 500,
                          compactness = 30, return_slic_data = TRUE, 
                          return_labels = TRUE, write_slic = "", verbose = FALSE)
imageShow(seg$slic_data)
```

#### 3. Perform an unsupervised classification
We will try to methods. The first one is k-means
```{r include = TRUE, message=F, warning=F, eval=FALSE}
# Extract pixel values and remove any NA
UnmixVal=getValues(unmixPCA)
i <- which(!is.na(UnmixVal))
UnmixVal <- na.omit(UnmixVal)

# perform a k-means classification
E <- kmeans(UnmixVal, 12, iter.max = 100, nstart = 10)
kmeans_raster <- raster(unmixPCA)
kmeans_raster[i] <- E$cluster
```

The second method combines k-means classification with unsupervised rando forest
```{r include = TRUE, message=F, warning=F, eval=FALSE}
# Perform a random forest classification with k-means
sampval<-UnmixVal[sample(nrow(UnmixVal), 500),]
rf_raster<- predict(unmixPCA,rf) #unsupervised rf
rf_prox <- randomForest(sampval,ntree = 1000, proximity = TRUE)$proximity

E_rf <- kmeans(rf_prox, 12, iter.max = 100, nstart = 10)
rf <- randomForest(sampval,as.factor(E_rf$cluster),ntree = 500)
rf_raster<- predict(unmixPCA,rf)
```

Let's plot both results using an optimized color palette
```{r include = TRUE, message=F, warning=F, eval=FALSE}
# produce optimal colors to plot the map
qual_col_pals = brewer.pal.info[brewer.pal.info$category == 'qual',]
col_vector = unlist(mapply(brewer.pal, qual_col_pals$maxcolors, rownames(qual_col_pals)))
colors=sample(col_vector, n)

# plot classifications with optimal colors and compare with the original image
plot(kmeans_raster, col=colors)
plot(rf_raster, col=colors)
plotRGB(unmixPCA, r=1, g=2, b=3, stretch="lin")
```

#### 4. Assign the majority class to all pixels in each segment
```{r include = TRUE, message=F, warning=F, eval=FALSE}
# Explore the structure of objects based on the original blog
str(unmixPCA)
str(rgb)
str(seg)

# sort the number of segments
sort(unique(as.vector(seg$labels)))

# Convert regions into raster
regionsrast=raster(seg$labels, unmixPCA[[1]])

# Assign to each segment the majority class in the unsupervised classification
rf_rasterVal=getValues(rf_raster)

# obtain the majority value for each segment
dataset=data.frame(cbind(rf_rasterVal, as.vector(t(seg$labels))))
names(dataset)=c('category', "item")

# Define a function to calculate majority per segment
g <- function(number) {
  ux <- unique(number)
  ux[which.max(tabulate(match(number, ux)))]
}

# Apply function to all segments and plot them
modes_by_cat <- summarise(group_by(dataset, item), group_mode = g(category))
out=inner_join(dataset, modes_by_cat)
val=out$group_mode
testrast=setValues(unmixPCA[[1]], val)
#testrast=setValues(imgCrop[[1]], testval)
plot(testrast, col=colors)
imageShow(seg$slic_data)
```

#### Homework

Download the image "PENDING" from the session 3 module in canvas: PENDING. Then adapt the code presented in sections ... in the lab instructions to ...

#### Sources
Here are some online sources consulted for developing this lab code:
https://www.r-bloggers.com/2020/03/analyzing-remote-sensing-data-using-image-segmentation/
http://remote-sensing.org/unsupervised-classification-with-r/ 
https://stackoverflow.com/questions/3823211/convert-a-matrix-to-a-1-dimensional-array