---
title: 'Advanced Remote Sensing Lab 3: Spatial segmentation, object-based analyses, and contextual metrics'
author: "Victor Gutierrez (victorhugo@temple.edu)"
output:
  html_document:
    df_print: paged
  word_document: default
  pdf_document: default
---

### Lab due
Febraury 16 2021

### Goal
1. To become familiar with textural analysis and image segmentation using the superpixels implementation in R.
2. To learn how to combine segmentation with pixel-bawesd classification to optimize classificatio results

### Total score
The lab counts for up to 3 points towards the final grade of the course.

Have fun!

```{r include = FALSE}
# Load screenshots
wd="//Users/tug61163/Documents/Courses/AdvancedRS/Spring2021/Session4_SegmentationTextural"
library(knitr)    # For knitting document and include_graphics function
library(ggplot2)  # For plotting
library(png)      # For grabbing the dimensions of png files
setwd(wd)
#fig1path=paste(wd, "SelEdit.png", sep="/")
#fig2path=paste(wd, "Sel2Edit.png", sep="/")
#fig3path=paste(wd, "Triangle.png", sep="/")
#fig4path=paste(wd, "Fig4.png", sep="/")
#fig1 = readPNG(fig1path)
#fig2 = readPNG(fig2path)
#fig3 = readPNG(fig3path)
#fig4 = readPNG(fig4path)
```

### Lab instructions

### 1. Environment setup and data loading
We will use a NAIP image (https://www.usgs.gov/centers/eros/science/usgs-eros-archive-aerial-photography-national-agriculture-imagery-program-naip?qt-science_center_objects=0#qt-science_center_objects) acquired on October 2019 near Mansfield, OH. You can download it from here: https://drive.google.com/file/d/15qXV3z6KBjRoCR-aABr1zrchPYi1ojiS/view?usp=sharing

Save downloaded image in a folder that you will use as a working directory. 

Setup working directory and load required libraries. In my case I saved it here: "/Users/tug61163/Documents/Courses/AdvancedRS/Spring2021/Session4_SegmentationTextural".
```{r include = TRUE, message=F, warning=F, eval=FALSE}
setwd("/Users/tug61163/Documents/Courses/AdvancedRS/Spring2021/Session4_SegmentationTextural")
library(raster)
library(OpenImageR)
library(randomForest)
library(RStoolbox)
library(dplyr)
library(RColorBrewer)
```

Unzip, stack and plot image. Then crop it to a manageable size
```{r include = TRUE, message=F, warning=F, eval=FALSE}
# Minessota
unzip("m_4309521_nw_15_1_20170920.zip")
img=stack("m_4309521_nw_15_1_20170920.tif")

# Ohio
unzip("m_4008215_ne_17_060_20191009.zip")
img=stack("m_4008215_ne_17_060_20191009.tif")

##l8unmix=stack("unmixPCA.tif")
plotRGB(img, r=3, g=2, b=1, stretch="lin")
plotRGB(img, r=4, g=3, b=2, stretch="lin")
e=drawExtent()
img=crop(img, e)
```

#### 2. Image segmentation
The package to be used for segmentation only recognizes images in JPEG format. Here we will convert the image from raster to JPEG format while conserving the original dimensionality.
```{r include = TRUE, message=F, warning=F, eval=FALSE}
nrows=img@nrows
ncols=img@ncols
jpeg("img.jpg", width = ncols, height = nrows)
plotRGB(img, r=4, g=3, b=2, stretch="lin")
dev.off()
```

We will use the super-pixels method. Modify the superpixel argument until you obtain satisfactory results. Larger values translate into more and smaller superpixels. You can learn more about the superpixels implementation here: https://infoscience.epfl.ch/record/177415 
```{r include = TRUE, message=F, warning=F, eval=FALSE}
rgb <- readImage("img.jpg")
seg = superpixels(input_image =rgb, method = "slic", superpixel = 1000,
                          compactness = 30, return_slic_data = TRUE, 
                          return_labels = TRUE, write_slic = "", verbose = FALSE)
imageShow(seg$slic_data)
```

#### 3. Feature extraction and classification
We will try two methods. The first one is k-means
```{r include = TRUE, message=F, warning=F, eval=FALSE}
# Extract pixel values and remove any NA
imgVal=getValues(img)
i <- which(!is.na(imgVal))
imgVal <- na.omit(imgVal)

# perform a k-means classification
E <- kmeans(imgVal, 12, iter.max = 100, nstart = 10)
kmeans_raster <- raster(img)
kmeans_raster[i] <- E$cluster
```

The second method combines k-means classification with unsupervised random forest
```{r include = TRUE, message=F, warning=F, eval=FALSE}
# Perform a random forest classification with k-means
sampval<-imgVal[sample(nrow(imgVal), 500),]
rf_prox <- randomForest(sampval,ntree = 1000, proximity = TRUE)$proximity

E_rf <- kmeans(rf_prox, 12, iter.max = 100, nstart = 10)
rf <- randomForest(sampval,as.factor(E_rf$cluster),ntree = 500)
rf_raster<- predict(img,rf)
```

Let's plot both results using an optimized random color palette
```{r include = TRUE, message=F, warning=F, eval=FALSE}
# produce optimal colors to plot the map
qual_col_pals = brewer.pal.info[brewer.pal.info$category == 'qual',]
col_vector = unlist(mapply(brewer.pal, qual_col_pals$maxcolors, rownames(qual_col_pals)))
colors=sample(col_vector, 12)

# plot classifications with optimal colors and compare with the original image
plot(kmeans_raster, col=colors)
plot(rf_raster, col=colors)
plotRGB(img, r=4, g=3, b=2, stretch="lin")
```

#### 4. Assign the majority class to all pixels in each segment
```{r include = TRUE, message=F, warning=F, eval=FALSE}
# Explore the structure of objects based on the original blog
str(img)
str(rgb)
str(seg)

# sort the number of segments
sort(unique(as.vector(seg$labels)))

# Convert regions into raster
regionsrast=raster(seg$labels, img[[4]])

# Assign to each segment the majority class in the unsupervised classification
rf_rasterVal=getValues(rf_raster)

# obtain the majority value for each segment
dataset=data.frame(cbind(rf_rasterVal, as.vector(t(seg$labels))))
names(dataset)=c('category', "item")

# Define a function to calculate majority per segment
g <- function(number) {
  ux <- unique(number)
  ux[which.max(tabulate(match(number, ux)))]
}

# Apply function to all segments and plot them
modes_by_cat <- summarise(group_by(dataset, item), group_mode = g(category))
out=inner_join(dataset, modes_by_cat)
val=out$group_mode
segclass=setValues(img[[1]], val)
#testrast=setValues(imgCrop[[1]], testval)
plot(segclass, col=colors)
```

#### Homework

Download the image for your homework form here: https://drive.google.com/file/d/1XSgrjKfWzv0lCh0_VbuV58wHq6DbFoh-/view?usp=sharing 

Then adapt the code presented in the lab to segment the image and produce an object based classification. Feel free to subset the image to any area of interest (between 1/3 and 1/4th of the total area of the image). Please work with different values for the arguments of the superpixels function related to compactness and superpixels and select the one that works best for the segmentation.

You can implement either the k-means or the hybrid k-means/random forest algorithm for the pixel-based classification. Fill up this form with the requested information: https://forms.office.com/Pages/ResponsePage.aspx?id=74FucSK1c0SOMRC9Asz25dmnkCS0Q29AsedCc0cCybpUM05HMUZWNVhGTkJTWEFUQVdKTk40UUpXSi4u  


#### Sources
Here are some online sources consulted for developing this lab code:
https://www.r-bloggers.com/2020/03/analyzing-remote-sensing-data-using-image-segmentation/
http://remote-sensing.org/unsupervised-classification-with-r/ 
https://stackoverflow.com/questions/3823211/convert-a-matrix-to-a-1-dimensional-array