---
title: 'Remote Sensing Lab 6: Image transformation and spectral indices
  in R'
author: "Victor Gutierrez (victorhugo@temple.edu)"
output:
  word_document: default
  pdf_document: default
  html_document:
    df_print: paged
always_allow_html: yes
---

```{r include = FALSE}
# Load screenshots
wd="/Users/Victor/Documents/Courses/IntroRemoteSensing/2019Spring/Class6/Lab6Materials"
setwd(wd)
library(knitr)    # For knitting document and include_graphics function
library(kableExtra)
knitr::opts_knit$set(root.dir = wd)
#knitr::knit(input=wd, output=wd)
```

## Lab due
March 13 2019

OJO: CREATE A FUNCTION TO RUN THE GRAPHS AND ONLY ASK THEM TO EVALUTE
THE COMPARISON IN ONE OF THEM (MAYBE TASSELED CAP)

## Goals
1. To learn the calculation and implementation of different image transformations and spectral indices.
2. To understand their interpretation and applicatio for land cover discrimination.

## Total score
The lab counts for up to 4 points towards the final grade of the course.

## Lab instructions
This lab assumes that some of the procedures learned in previous labs for downloading and manipulating images in R are already mastered and therefore are not explained in detail here. Refer to previous labs if there is any problem with those procedures.

1. Go to earthexplorer.usgs.gov. Enter “Pucallpa” as the place in the search criteria. In datasets select Landsat > Landsat Collection-1 Level-1 > Landsat 8 OLI/TIRS C1 Level-1.

2. In results, select Landsat as the dataset and download image corresponding to path 06, row  066 from August 04-2016.  Download the GEOTIFF data product.

3. Open R Studio, open a new R script and change the route to your working directory in the script using the function setwd().

4. Copy and paste each chunk of code in your new R script and run it trying to understand the purpose, logic and syntaxis of each line. Make sure the code runs with no errors before moving to the next one.

5.	Answer the questions in the answer sheet and submit it along with the required pdf file to canvas.

Have fun!

1. Load required libraries.
```{r include = TRUE, message=F, warning=F}
library(raster)
library(rgdal)
library(RStoolbox)
```

2. Set working directory: Change the path below by copying and pasting the route to any selected folder in your workstation. If you are using a windows machine, make sure you use forward slash instead of backward
```{r include = TRUE, message=F, warning=F}
wd="/Users/Victor/Documents/Courses/IntroRemoteSensing/2019Spring/Class6/Lab6Materials"
#dir.create(wd)
setwd(wd)
```

3. Decompress, open and stack landsat image
```{r include = TRUE, message=F, warning=F}
dir()

L8name="LC08_L1TP_006066_20160804_20170322_01_T1"

# Decompress and stack landsat image
untar(paste(L8name, "tar", sep="."))

meta=readMeta(paste(paste(L8name, "MTL", sep="_"), "txt", sep="."))
L8=stackMeta(paste(paste(L8name, "MTL", sep="_"), "txt", sep="."))
L8
plotRGB(L8, r=5, g=4, b=3, stretch="lin")
```

4. Use the drawExtent() function to create an extent that includes approximately the area shown in the plot below. Use that extent to crop the Landsat scene.
```{r include = TRUE, message=F, warning=F}
#e=drawExtent()
e=extent(513128.8, 599389.9 , -1001877, -894184.4)
L8rsz=crop(L8, e)
plotRGB(L8rsz, r=5, g=4, b=3, stretch="lin")

# Remove objects to release memory
rm(L8)
```

6. Convert pixel values from digital numbers to ground reflectance
```{r include = TRUE, message=F, warning=F}
haze=estimateHaze(L8rsz, hazeBands=1:5, 
                  darkProp=0.001, plot=TRUE)

L8dos=radCor(L8rsz, metaData=meta, 
             method="sdos", hazeValues=haze, hazeBands=1:5)

L8subset=L8dos[[2:7]]
rm(L8dos)
rm(L8rsz)
```

7. Calculate and plot band transformations and spectral indices
```{r include = TRUE, message=F, warning=F}
L8PCA=rasterPCA(L8subset, nComp=nlayers(L8subset)) # Principal components analysis

# Plot each band at a time
plot(L8PCA$map[[1]], col = gray.colors(100, start = 0.1, end = 0.9, gamma = 2.2, alpha = NULL), axes=FALSE)

plotRGB(L8PCA$map, r=1, g=2, b=3, stretch="lin")
plotRGB(L8PCA$map, r=4, g=5, b=6, stretch="lin")

L8indices=spectralIndices(L8subset, blue=1, green=2,
                          red=3, nir=4, swir2=5, swir3=6,
                          indices=c("NDVI",  "NDWI", "NBRI")) # Spectral indices
plotRGB(L8indices, r=1, g=2, b=3, stretch="lin")

L8tascap=tasseledCap(L8subset, sat="Landsat8OLI") # Tasseled cap
plotRGB(L8tascap, r=1, g=2, b=3, stretch="lin")

L8allbands=stack(L8subset, L8indices, L8tascap)
rm(L8subset, L8indices, L8tascap, L8PCA, haze)
```

8. Draw different extents and collect the spectral information of a pixel that best typifies the land covers referenced below

```{r include = TRUE, message=F, warning=F}
# plotRGB(L8allbands, r=10, g=11, b=12, stretch="lin")
# urbanext=drawExtent() # draw a different extent for each land cover type
# plotRGB(L8allbands, r=10, g=11, b=12, stretch="lin", ext=urbanext)
# urban=click(L8allbands, n=1, cell=TRUE)
# 
# plotRGB(L8allbands, r=10, g=11, b=12, stretch="lin")
# forext=drawExtent() # draw a different extent for each land cover type
# plotRGB(L8allbands, r=10, g=11, b=12, stretch="lin", ext=forext)
# forest=click(L8allbands, n=1, cell=TRUE)
# 
# plotRGB(L8allbands, r=10, g=11, b=12, stretch="lin")
# lakext=drawExtent() # draw a different extent for each land cover type
# plotRGB(L8allbands, r=10, g=11, b=12, stretch="lin", ext=lakext)
# lake=click(L8allbands, n=1, cell=TRUE)
# 
# plotRGB(L8allbands, r=10, g=11, b=12, stretch="lin")
# rivext=drawExtent() # draw a different extent for each land cover type
# plotRGB(L8allbands, r=10, g=11, b=12, stretch="lin", ext=rivext)
# river=click(L8allbands, n=1, cell=TRUE)
# 
# plotRGB(L8allbands, r=10, g=11, b=12, stretch="lin")
# grassext=drawExtent() # draw a different extent for each land cover type
# plotRGB(L8allbands, r=10, g=11, b=12, stretch="lin", ext=grassext)
# grass=click(L8allbands, n=1, cell=TRUE)
# 
# plotRGB(L8allbands, r=10, g=11, b=12, stretch="lin")
# burnext=drawExtent() # draw a different extent for each land cover type
# plotRGB(L8allbands, r=10, g=11, b=12, stretch="lin", ext=burnext)
# burnt=click(L8allbands, n=1, cell=TRUE)
# 
# plotRGB(L8allbands, r=10, g=11, b=12, stretch="lin")
# sandext=drawExtent() # draw a different extent for each land cover type
# plotRGB(L8allbands, r=10, g=11, b=12, stretch="lin", ext=sandext)
# sand=click(L8allbands, n=1, cell=TRUE)
# 
# plotRGB(L8allbands, r=10, g=11, b=12, stretch="lin")
# oilext=drawExtent() # draw a different extent for each land cover type
# plotRGB(L8allbands, r=10, g=11, b=12, stretch="lin", ext=oilext)
# oilpalm=click(L8allbands, n=1, cell=TRUE)
# 
# landcovers=data.frame(rbind(urban, forest, lake, river,
#                             grass, burnt, sand, oilpalm))
# 
# rm(urban, forest, lake, river, grass, burnt, sand, oilpalm)
# landcovers=as.data.frame(t(as.matrix(landcovers)))
# landcovers=landcovers[2:nrow(landcovers),]
# 
# names(landcovers)=c('urban', 'forest', 'lake', 'river',
#                     'grass', 'burnt', 'sand', 'oilpalm')
# 
# colors=c('red', 'green','blue', 'gray', 'yellow', 'black', 'purple', 'orange')
# 
# graphics.off() # removes all plots from the environment
```

9. Create an aggregated pdf file with 
a) RGB pixel maps representing  the zoomed in insets used to collect pixel values per land covers and 
b) three line plots representing pixel values per land cover along different bands using ground reflectance, spectral indices, and a tasseled cap transformation.
```{r include = TRUE, message=F, warning=F}
#pdf("LCplot.pdf")
######### PLOT EXTENTS USED FOR COLLECTING PIXEL VALUES
# plotRGB(L8allbands, r=10, g=11, b=12, stretch="lin",ext=urbanext, axes=TRUE, main="Urban")
# plotRGB(L8allbands, r=10, g=11, b=12, stretch="lin", ext=forext, axes=TRUE, main="Forest") 
# plotRGB(L8allbands, r=10, g=11, b=12, stretch="lin", ext=lakext, axes=TRUE, main="Lake")
# plotRGB(L8allbands, r=10, g=11, b=12, stretch="lin", ext=rivext, axes=TRUE, main="River")
# plotRGB(L8allbands, r=10, g=11, b=12, stretch="lin", ext=grassext, axes=TRUE, main="Grass")
# plotRGB(L8allbands, r=10, g=11, b=12, stretch="lin", ext=burnext, axes=TRUE, main="Burned")
# plotRGB(L8allbands, r=10, g=11, b=12, stretch="lin", ext=sandext, axes=TRUE, main="Sand")
# plotRGB(L8allbands, r=10, g=11, b=12, stretch="lin", ext=oilext, axes=TRUE, main="Oil palm")

######### PLOT SPECTRAL RESPONSES GRAPHS
# Names of the labels to add to the x axis in the plot
# labnames=c("B2", "B3", "B4", "B5", "B6", "B7", 
#                     "NDVI","NDWI", "NBRI", 
#                     "bright", "green", "wet")
# 
# # PLOT REFLECTANCE BANDS
# minband=1 # minimum band to plot
# maxband=6 # maximum band to plot
# 
# # Add extra space to right of plot area; change clipping to figure
# par(mar=c(5.1, 4.1, 4.1, 8.1), xpd=TRUE)
# plot(landcovers[minband:maxband, 1], type="l", col="red",
#      ylab="pixel value", xlab="band", 
#      ylim=c(min(landcovers[minband:maxband,]),
#      max(landcovers[minband:maxband,])), xaxt="n", 
#      main="Band reflectance per land cover")
# lines(landcovers[minband:maxband, 2], type="l", col="green")
# lines(landcovers[minband:maxband, 3], type="l", col="blue")
# lines(landcovers[minband:maxband, 4], type="l", col="pink")
# lines(landcovers[minband:maxband, 5], type="l", col="yellow")
# lines(landcovers[minband:maxband, 6], type="l", col="gray")
# lines(landcovers[minband:maxband, 7], type="l", col="orange")
# lines(landcovers[minband:maxband, 8], type="l", col="black")
# lines(landcovers[minband:maxband, 9], type="l", col="purple")

# legend("topright", inset=c(-0.3, 0), legend=names(landcovers),
#        col=c("red", "green", "blue", "pink", "yellow",
#              "gray", "orange", "black", "purple"), lty=1, cex=0.8)
# dev.off()
```

## Lab 6 deliverables

1. Respond the questions below based on the colors of different land covers after applying the tasseledCap() function (0.6): 
a. What land cover results in the redest colors?  ___________________________ why? ___________________________________________________________________

b. What land cover results in the bluest colors?  ____________________________
why? __________________________________________________________________

c. What land cover results in the greenest color (you might have to plot this band as a grey scale to identify the brightest pixels)?  _________________________________
why? __________________________________________________


Check the plots representing pixel values per different land covers for different bands that appear in your pdf file and answer the questions below:

2. If you had to select one band in each of the plots representing pixel values per land cover (0.6):
a. provide the name of the band that would provide the best discrimination between land covers in each plot
* band reflectance per land cover: ________________
* vegetation index value per land cover ________________
* Tasseled cap value per land cover _________________

b. provide the name of the band that would provide the worst discrimination between land covers?
* band reflectance per land cover: ________________
* vegetation index value per land cover ________________
* Tasseled cap value per land cover _________________

3. What two land covers do you expect to have the best discrimination in each plot type (0.3)?
* band reflectance per land cover: ________________
* vegetation index value per land cover ________________
* Tasseled cap value per land cover _________________

4. What two land covers do you expect to have the worst discrimination in each plot type (0.3)?
* band reflectance per land cover: ________________
* vegetation index value per land cover ________________
* Tasseled cap value per land cover _________________

5. Briefly explain your rationale to your answers in 2 to 4 above (0.2): _________________________________________

To upload in canvas:
1.	All the figures generated in one single pdf file (2 pt)
2.	This document with your answers to the questions above (2 pt).
