---
title: 'Remote Sensing Lab 5: Topographic correction
  in R'
author: "Victor Gutierrez (victorhugo@temple.edu)"
output:
  word_document: default
  pdf_document: default
  html_document:
    df_print: paged
always_allow_html: yes
---

```{r include = FALSE}
# Load screenshots
wd="/Users/Victor/Documents/Courses/IntroRemoteSensing/2019Spring/Class5/Lab5Materials"
setwd(wd)
library(knitr)    # For knitting document and include_graphics function
library(kableExtra)
knitr::opts_knit$set(root.dir = wd)
#knitr::knit(input=wd, output=wd)
```

## Lab due
February 20 2019

## Goals

1.	To learn how to align and mask images with different geographic projections and spatial resolutions.
2.	To master methods for topographic adjustment to spectral satellite images
3.	To understand the purpose of and methods for alligning raster files (projecting, cropping and masking).
4.	To learn how to mitigate the effects of illumination effets in the spectral characteristics of satellite imagery from mountainous areas. 

## Total score
The lab counts for up to 4 points towards the final grade of the course.

## Lab instructions
This lab assumes that some of the procedures learned in previous labs for downloading and manipulating images in R are already mastered and therefore are not explained in detail here. Refer to previous labs if there is any problem with those procedures.

1. Go to earthexplorer.usgs.gov. Enter “Hanzhong” as the place in the search criteria. In datasets select:
•	Landsat > Landsat Collection-1 Level-1 > Landsat 8 OLI/TIRS C1 Level-1.
•	Digital Elevation > SRTM > SRTM 1 Arc-Second Global.

2. In results, select Landsat as the dataset and download image corresponding to path 128, row  37 from 13-MAY-18.  Download the GEOTIFF data product.

3. Select SRTM as the dataset and download the only image available. It should correspond to the  coordinates 33, 107. Download the GEOTIFF product.

4. Open R Studio, open a new R script and change the route to your working directory in the script using the function setwd().

3. Copy and paste each chunk of code in your new R script and run it trying to understand the purpose, logic and syntaxis of each line. Make sure the code runs with no errors before moving to the next one.

5.	If the code runs, adapt the script to apply it to the “Rocky Mountains”. The Landsat image corresponds to path 38 row 29 from 10-AUG-18. The SRTM image correspond to the coordinates 44, -110.

6. Answer the questions in the answer sheet and submit it along with the required pdf file to canvas.

Have fun!

1. Load required libraries.
```{r include = TRUE, message=F, warning=F}
library(raster)
library(rgdal)
library(RStoolbox)
```

2. Set working directory: Change the path below by copying and pasting the route to any selected folder in your workstation. If you are using a windows machine, make sure you use forward slash instead of backward
```{r include = TRUE, message=F, warning=F}
wd="/Users/Victor/Documents/Courses/IntroRemoteSensing/2019Spring/Class5/Lab5Materials"
setwd(wd)
```

3. Decompress, open and stack landsat image
```{r include = TRUE, message=F, warning=F}
dir()

### IN IMAGE FOR CLASS EXCERCISE
#L8Name="LC08_L1TP_128037_20180513_20180517_01_T1"

### IN IMAGE FOR LAB REPORT
L8Name="LC08_L1TP_038029_20180810_20180815_01_T1"

# Decompress and stack landsat image
untar(paste(L8Name, "tar", sep="."))

meta=readMeta(paste(paste(L8Name, "MTL", sep="_"), "txt", sep="."))
L8=stackMeta(paste(paste(L8Name, "MTL", sep="_"), "txt", sep="."))
L8
plot(L8[[5]])
```

4. Open and project SRTM image. Then plot it.
```{r include = TRUE, message=F, warning=F}
### IN IMAGE FOR CLASS EXCERCISE
#srtmName="n33_e107_1arc_v3.tif"

### IN IMAGE FOR LAB REPORT
srtmName="n44_w110_1arc_v3.tif"

srtm=raster(srtmName)      
srtm
plot(srtm)

srtmPrj=projectRaster(srtm, L8)
plot(srtmPrj)
```

5. Use the drawExtent() function to create an extent that includes only pixels with valid information in the plotted SRTM image. Use that extent to crop both the projected srtm image and the Landsat scene.
```{r include = TRUE, message=F, warning=F}
#e=drawExtent()
#e=extent(686932.5, 778498, 3656499, 3763326) # For class example
e=extent(580441.2, 654659.8,4873903, 4982024) # for lab report
srtmRsz=crop(srtmPrj,e)
plot(srtmRsz)

L8Rsz=crop(L8, e)
plot(L8Rsz[[5]])

# Remove objects to release memory
#rm(L8,srtm, srtmPrj)
```

6.  Mask images to a common valid domain
```{r include = TRUE, message=F, warning=F}
m=c(-Inf,0,NA, 0,Inf,1)
reclassmatrix=matrix(m, ncol=3, byrow=TRUE)
L8msk=reclassify(L8Rsz, reclassmatrix)
L8msk=Reduce(f="*", as.list(L8msk))
srtmsk=reclassify(srtmRsz, reclassmatrix)

mask=L8msk*srtmsk
plot(mask)

L8mskd=mask(L8Rsz,mask)
srtmskd=mask(srtmRsz, mask)

plot(L8mskd[[5]])
plot(srtmskd)

# Remove elements to release memory
#rm(mask, L8msk, L8msk,srtmsk, L8Rsz, srtmRsz)
```

7. Calculate terrain variables from the DEM
```{r include = TRUE, message=F, warning=F}
slope=terrain(srtmskd, opt='slope')
aspect=terrain(srtmskd, opt='aspect')
terrainvar=stack(slope,aspect)
names(terrainvar)=c('slope', 'aspect')
rm(slope,aspect)

# calculate illumination
shade=topCor(L8mskd, dem=terrainvar, metaData=meta, method="illu")
plot(shade)
```

8.  Apply topographic correction
```{r include = TRUE, message=F, warning=F}
L8cor=topCor(L8mskd, dem=terrainvar, metaData=meta, method="C")

pdf("VGutierrez_Lab5.pdf")
  plotRGB(L8mskd, r=5, g=4, b=3, axes=FALSE, stretch="lin")
  plotRGB(L8cor, r=5, g=4, b=3, axes=FALSE, stretch="lin")
dev.off()
```

## Lab 5 deliverables

1. Upload two RGB map composites representing the pixel values for bands 543 from Landsat before and after topographic correction for the Rocky mountains. Change the name of the pdf file for your Last name First initial and the name of the lab assignment ("GutierrezV_Lab1.pdf"") before submission (2 pts).

2. Read the help documentation for the function topCor() and then answer the questions below briefly (0.5 pts each):

In the application of the topCor function to produce an illumination model:
a. What are the two variables that it calculates from the DEM raster?

b. What are the two variables that it obtains from the Landsat metadata file? What is the value of those two variables in the Landsat scene used for the Rocky mountains?

c. What does the application of the “illu” method in the function topCor() produce?

d. What does the application of the method “C” in topCor() produce (check the images in the BeforeAndAfter.pdf file obtained from running the algorithm)