---
title: 'Applied Machine Learning for Spatial Analysis. Lab Week 6: Exploratory Data Analysis (EDA)'
author: "Victor Gutierrez (victorhugo@temple.edu)"
output:
  html_document:
    df_print: paged
  pdf_document: default
  word_document: default
---

```{r include = FALSE}
# Load screenshots
wd="/Users/tug61163/Library/CloudStorage/OneDrive-TempleUniversity/Repositories/Gitrepo/MALESA_Code"
setwd(wd)
library(knitr)    # For knitting document and include_graphics function
library(ggplot2)  # For plotting
library(png)      # For grabbing the dimensions of png files

fig1path=paste(wd, "Fig1.png", sep="/")
#fig2path=paste(wd, "Sel2Edit.png", sep="/")
#fig3path=paste(wd, "Triangle.png", sep="/")
#fig4path=paste(wd, "Fig4.png", sep="/")
#fig1 = readPNG(fig1path)
#fig2 = readPNG(fig2path)
#fig3 = readPNG(fig3path)
#fig4 = readPNG(fig4path)
```

## Goal
To learn how to perform exploratory data analysis.

## Total score
The lab counts for up to 4 points towards the final grade of the course.

## Lab instructions
1.	Launch R Studio and open a new R script: File/ New File/ R Script. Then save it as a new file: File/ Save As…
2.	Read the instructions below step by step. Copy and paste each chunk of code at a time in your R script. Select the code with your mouse or shift/arrow keys and then run it by pressing the keys control-enter simultaneously.

## Lab overview
For this lab, we are going to use the mosquito data that we 

Have fun!

### 1. System setup
```{r include = TRUE, message=F, warning=F, eval=FALSE}
wd=("/Users/tug61163/Library/CloudStorage/OneDrive-TempleUniversity/Courses/MALESA/2022Fall/Week6")
setwd(wd)
library(terra)
library(sf)
library(DataExplorer)
library(GGally)
library(dplyr)
```

#### Load files
```{r include = TRUE, message=F, warning=F, eval=FALSE}
load("alldata.RData")
```

#### Preparing the data
Before exploring the data, let's create a binary column representing mosquito presences/absences.
```{r include = TRUE, message=F, warning=F, eval=FALSE}
alldata$PRESENCE=alldata$ADULTS
alldata$PRESENCE[alldata$PRESENCE>0]=1
alldata$PRESENCE=as.factor(alldata$PRESENCE)
```

#### Identify the structure of the data
```{r include = TRUE, message=F, warning=F, eval=FALSE}
str(alldata)
plot_str(alldata)
summary(alldata)
head(alldata)
```

#### Identify and remove observations with missing values and data duplicates
```{r include = TRUE, message=F, warning=F, eval=FALSE}
nrow(alldata)
alldata=subset(alldata, complete.cases(alldata))
nrow(alldata)

# Identify and remove duplicates
dups=duplicated(alldata)
alldata[dups,]
nrow(alldata)
alldata=alldata[!dups,]
nrow(alldata)
```

#### Explore the characteristics of individual variables
```{r include = TRUE, message=F, warning=F, eval=FALSE}
names(alldata)
datasubset=alldata[,c(2, 6:10, 13:33)]
plot_bar(datasubset)
plot_qq(datasubset)
plot_density(datasubset)
plot_prcomp(datasubset[,c(2,3, 7:24)], nrow=2L, ncol=2L)

```

#### Check interactions between pairs of data 
```{r include = TRUE, message=F, warning=F, eval=FALSE}
cormat=cor(datasubset[,c(2, 3, 7:20, 22:26)], use="complete.obs", method="pearson")
plot_correlation(datasubset)

pairs(datasubset[,c(7:13)])

panel.hist <- function(x, ...) {
    usr <- par("usr")
    on.exit(par(usr))
    par(usr = c(usr[1:2], 0, 1.5))
    his <- hist(x, plot = FALSE)
    breaks <- his$breaks
    nB <- length(breaks)
    y <- his$counts
    y <- y/max(y)
    rect(breaks[-nB], 0, breaks[-1], y, col = rgb(0, 1, 1, alpha = 0.5), ...)
    # lines(density(x), col = 2, lwd = 2) # Uncomment to add density lines
}
pairs(datasubset[,c(7:13)], upper.panel = NULL,
      diag.panel = panel.hist)
plot_bar(datasubset, by="PRESENCE")
```

#### Plot data interactions per groups
```{r include = TRUE, message=F, warning=F, eval=FALSE}
options(repr.plot.width = 20, repr.plot.height = 10)
alldata %>% 
  select("roughness", "tmea", "income", "latino", "TRI") %>%
  ggpairs(mapping = aes(color = alldata$PRESENCE, alpha = 0.5))

boxplot(income~PRESENCE,data=alldata, main="income per mosquito presence", 
   xlab="presence/absence", ylab="income")

boxplot(TRI~PRESENCE,data=alldata, main="TRI per mosquito presence", 
   xlab="presence/absence", ylab="TRI")

pairs(datasubset[,c(7:13)],                     # Data frame of variables
      labels = colnames(datasubset[,c(7:13)]),  # Variable names
      pch = 21,                                 # Pch symbol
      bg = rainbow(3)[datasubset$PRESENCE],     # Background color of the symbol (pch 21 to 25)
      col = rainbow(3)[datasubset$PRESENCE],    # Border color of the symbol
      main = "Iris dataset",                    # Title of the plot
      row1attop = TRUE,                         # If FALSE, changes the direction of the diagonal
      gap = 1,                                  # Distance between subplots
      cex.labels = NULL,                        # Size of the diagonal text
      font.labels = 1)                          # Font style of the diagonal text
```

#### Produce report
Let's produce a report using the package "DataExplorer"
```{r include = TRUE, message=F, warning=F, eval=FALSE}
alldata %>%
    create_report(
      #output_file = paste("Report", format(Sys.time(), "%Y-%m-%d %H:%M:%S %Z"), sep=" - "),
       # Some people reported a png() print error. So I chabged the line above for this one:
      output_file = paste("Report"),
        report_title = "EDA Report - Mosquito presence",
        y = "PRESENCE"
    )
```

