---
title: 'Applied Machine Learning for Spatial Analysis. Lab Week 4: Bias mitigaton and area of applicabilty'
author: "Victor Gutierrez (victorhugo@temple.edu)"
output:
  html_document:
    df_print: paged
  pdf_document: default
  word_document: default
---

```{r include = FALSE}
# Load screenshots
wd="/Users/tug61163/Library/CloudStorage/OneDrive-TempleUniversity/Repositories/Gitrepo/MALESA_Code"
setwd(wd)
library(knitr)    # For knitting document and include_graphics function
library(ggplot2)  # For plotting
library(png)      # For grabbing the dimensions of png files

fig1path=paste(wd, "Fig1.png", sep="/")
#fig2path=paste(wd, "Sel2Edit.png", sep="/")
#fig3path=paste(wd, "Triangle.png", sep="/")
#fig4path=paste(wd, "Fig4.png", sep="/")
#fig1 = readPNG(fig1path)
#fig2 = readPNG(fig2path)
#fig3 = readPNG(fig3path)
#fig4 = readPNG(fig4path)
```
## Lab due
This lab has no report

## Goal
To learn how to detect and mitigate spatial biases in input data

## Lab instructions
1.	Launch R Studio and open a new R script: File/ New File/ R Script. Then save it as a new file: File/ Save As…
2.	Read the instructions below step by step. Copy and paste each chunk of code at a time in your R script. Select the code with your mouse or shift/arrow keys and then run it by pressing the keys control-enter simultaneously.

## Lab overview
For this lab, we are going to use the mosquito data from previous labs to identify, measure, and mitigate spatial biases in training data.

Have fun!

### 1. System setup
```{r include = TRUE, message=F, warning=F, eval=FALSE}
wd="/Users/tug61163/Library/CloudStorage/OneDrive-TempleUniversity/Courses/MALESA/2022Fall/Week13_SpatialElementsOfMachineLearning"
sourcedir="/Users/tug61163/Library/CloudStorage/OneDrive-TempleUniversity/Courses/MALESA/2022Fall/Week6"
setwd(wd)
library(terra)
library(sf)
#library(DataExplorer)
#library(GGally)
#library(dplyr)
```

#### Load files
```{r include = TRUE, message=F, warning=F, eval=FALSE}
load(paste0(sourcedir,"/alldata.RData"))
names(alldata)
summary(alldata$ADULTS)
```

#### Preparing the data
Before exploring the data, let's create a binary column representing mosquito presences/absences.
```{r include = TRUE, message=F, warning=F, eval=FALSE}
alldata$PRESENCE=alldata$ADULTS
alldata$PRESENCE[alldata$PRESENCE>0]=1
alldata$PRESENCE=as.factor(alldata$PRESENCE)
```

Identify and remove missing observations and duplicates
```{r include = TRUE, message=F, warning=F, eval=FALSE}
str(alldata)
summary(alldata)
head(alldata)
nrow(alldata)
alldata=subset(alldata, complete.cases(alldata))
nrow(alldata)

dups=duplicated(alldata)
alldata[dups,]
nrow(alldata)
alldata=alldata[!dups,]
nrow(alldata)
```

#### Assess and mitigate spatial bias
Upload raster stack and spatial sampled data points
```{r include = TRUE, message=F, warning=F, eval=FALSE}
datastack=rast(paste0(sourcedir, "/datastack.tif"))

# Remove NAs
sumstack=sum(datastack)
sumstack[!is.na(sumstack)]=1
datastack=mask(datastack, sumstack)

SpatData=vect(paste0(sourcedir,"/SpatData.shp"))
names(datastack)==names(alldata[13:32])
names(datastack[[15]])="LC"
```

Produce a grid to perform spatial filtering
```{r include = TRUE, message=F, warning=F, eval=FALSE}
r <- datastack[[16]]
# set the resolution of the cells to fifty times the original resolution
r=aggregate(r,fact=33) # approximately 1 km
r[(r>0)]=c(1:ncell(r))
plot(r)
r=as.polygons(r)
```

Extract from each grid-cell, a random sampled point
```{r include = TRUE, message=F, warning=F, eval=FALSE}
strat=extract(r, SpatData)
SpatData$strat=strat$income
for(i in unique(SpatData$strat)){
  sub=SpatData[SpatData$strat==i]
  samp=sample(sub, size=1)
  if(i==unique(SpatData$strat)[1]){
    subset=samp
  }else{
      subset=rbind(subset, samp)}
}
plot(r)
plot(SpatData, cex=0.3, add=T)
plot(subset, cex=0.3, col="red", add=T)
```

Compare the frequency distribution of covariates within the sampled locations to the one that would be obtained if sampling was performed using a random reference sampling
```{r include = TRUE, message=F, warning=F, eval=FALSE}
# Before filtering
reforig=spatSample(datastack[[16]], size=nrow(SpatData), "random", as.points=T)
length(reforig)
reforig=reforig[!is.na(reforig$income)]
length(reforig)
plot(datastack[[16]])
plot(reforig,add=T)

cdf1=ecdf(alldata$income)
cdf2=ecdf(reforig$income)
plot(cdf1)
plot(cdf2, col="red", add=T)

# After filtering
reffilt=spatSample(datastack[[16]], size=nrow(subset), "random", as.points=T)
reffilt=reffilt[!is.na(reffilt$income)]
subdata=extract(datastack, subset)
cdf1f=ecdf(subdata$income)
cdf2f=ecdf(reffilt$income)
plot(cdf1f, cex=0.05)
plot(cdf2f, cex=0.05, col="red", add=T)
```

It looks like it hasn't changed that much or that it might be worse after filtering, especially in higher income areas. 

#### Area of applicability
We will have to constraint the geographic domain fo the data to include only the samples within urban areas. For that purpose, we will download a map representing urban areas only from the US census: https://www.census.gov/cgi-bin/geo/shapefiles/index.php 

```{r include = TRUE, message=F, warning=F, eval=FALSE}
urban=vect(paste0(sourcedir, "/tl_2015_us_uac10.shp"))
urban=project(urban, datastack)
urban=crop(urban, ext(datastack))
#writeVector(urban, "urban.shp")
urbanrast=rasterize(urban, datastack)
datastackurb=mask(datastack, urbanrast)
urbanrast=mask(urbanrast, sumstack)
urban=as.polygons(urbanrast)
plot(urbanrast, col= "gray")
plot(subset, cex=0.5, add=T)
```


```{r include = TRUE, message=F, warning=F, eval=FALSE}
urbansf=sf::st_as_sf(urban)
subset=sf::st_as_sf(subset)
#urbansf=read_sf("urban.shp")
ptsInUrban=st_intersection(subset, urbansf)
ptsInUrban=vect(ptsInUrban)
plot(datastackurb[[16]])
plot(ptsInUrban,cex=0.2,add=T)

names(ptsInUrban)
ptsInUrban=ptsInUrban[,1:10]
```

Let's compare the frequency distribution of covariates between sampled and random points but only within urban areas.
```{r include = TRUE, message=F, warning=F, eval=FALSE}
set.seed(13)
reffiltU=spatSample(urban, size=nrow(ptsInUrban), "random")
varU=extract(datastack, reffiltU)
reffiltU$income=varU$income
#reffiltU=spatSample(datastackurb[[16]], size=nrow(ptsInUrban), "random", as.points=T)
#reffiltU=reffilt[!is.na(reffiltU$income)]

plot(datastackurb[[16]])
plot(ptsInUrban, cex=0.3, add=T)
plot(reffiltU, cex=0.3, add=T, col="red")

subdataU=extract(datastack, ptsInUrban)
cdf1U=ecdf(subdataU$income)
cdf2U=ecdf(reffiltU$income)
plot(cdf1U, cex=0.05)
plot(cdf2U, col="red", cex=0.05, add=T)

# Let's compare it with the previous one
plot(cdf1f, cex=0.05)
plot(cdf2f, cex=0.05, col="red", add=T)

# Through a k-s test
ks.test(subdataU$income, reffiltU$income)
ks.test(subdata$income, reffilt$income)
```

