---
title: 'Applied Machine Learning for Spatial Analysis. Lab Week 10: Advanced algorithms (supervised classification)'
author: "Victor Gutierrez (victorhugo@temple.edu)"
output:
  html_document:
    df_print: paged
  pdf_document: default
  word_document: default
---

```{r include = FALSE}
# Load screenshots
wd="/Users/tug61163/Library/CloudStorage/OneDrive-TempleUniversity/Repositories/Gitrepo/MALESA_Code"
setwd(wd)
library(knitr)    # For knitting document and include_graphics function
library(ggplot2)  # For plotting
library(png)      # For grabbing the dimensions of png files

fig1path=paste(wd, "Fig1.png", sep="/")
#fig2path=paste(wd, "Sel2Edit.png", sep="/")
#fig3path=paste(wd, "Triangle.png", sep="/")
#fig4path=paste(wd, "Fig4.png", sep="/")
#fig1 = readPNG(fig1path)
#fig2 = readPNG(fig2path)
#fig3 = readPNG(fig3path)
#fig4 = readPNG(fig4path)
```

## Goals
To learn how to tune and validate a variety of supervised machine learning models.

## Lab instructions
1.	Launch R Studio and open a new R script: File/ New File/ R Script. Then save it as a new file: File/ Save As…
2.	Read the instructions below step by step. Copy and paste each chunk of code at a time in your R script. Select the code with your mouse or shift/arrow keys and then run it by pressing the keys control-enter simultaneously.

## Lab overview
For this lab, we are going to use data from previous labs to calibrate and evaluated an array of ML algorithms and use them to produce spatial predictions.

The code described below was adapted from the following sources:
https://gist.github.com/hakimabdi/720f1481af9eca0b7b97d9856052e0e2 
https://stackoverflow.com/questions/21662180/using-neuralnet-with-caret-train-and-adjusting-the-parameters 

Have fun!

### System setup
```{r include = TRUE, message=F, warning=F, eval=FALSE}
wd=("/Users/tug61163/Library/CloudStorage/OneDrive-TempleUniversity/Courses/MALESA/2022Fall/Week12_AdvancedAlgorithms")
setwd(wd)

library(caret)
library(terra)
library(SpatialML)
library(sf)
#library(pdp)
sourcedir="/Users/tug61163/Library/CloudStorage/OneDrive-TempleUniversity/Courses/MALESA/2022Fall/Week9"
```

Load required files
```{r include = TRUE, message=F, warning=F, eval=FALSE}
load(paste0(sourcedir,"/sampdata.RData"))
refdata=sampdata[,-c(1, 18:19)]
```

### Cross validation setup

Partition data for calibration and validation
```{r include = TRUE, message=F, warning=F, eval=FALSE}
parts = createDataPartition(refdata$veg, p = .7, list = F)
train = refdata[parts, ]
test = refdata[-parts, ]
```

Standardize the data (required for neural networks, svm and others)
```{r include = TRUE, message=F, warning=F, eval=FALSE}
zscore=function(x){
  zval=(x-mean(x, na.rm=TRUE))/sd(x, na.rm=TRUE)
  return(zval)
}

trainx=data.frame(apply(train[1:(ncol(train)-1)], 2, FUN=zscore))
trainx=cbind(trainx, train[,ncol(train)])
names(trainx)[ncol(trainx)]="veg"
testx=data.frame(apply(test[1:(ncol(test)-1)], 2, FUN=zscore))
testx=cbind(testx, test[,ncol(test)])
names(testx)[ncol(testx)]="veg"
```

Setup the parameters for the k-fold cross-validation that will be passed to the train() function.
```{r include = TRUE, message=F, warning=F, eval=FALSE}
tc <- trainControl(method = "repeatedcv", #  cross-validation method
                   number = 5, # number of folds
                   repeats = 3, # number of repeats
                   allowParallel = TRUE, # allow multicore use
                   verboseIter = TRUE) # view the training iterations
```

### Modeling
Use the information associated all models available in caretto select the models to be used
```{r include = TRUE, message=F, warning=F, eval=FALSE}
names(getModelInfo())
gnni=getModelInfo("nnet")
gnni$nnet$type
gnni$nnet$library
modelLookup("nnet")
```

Specify grid for tunning each model
```{r include = TRUE, message=F, warning=F, eval=FALSE}
# Random forest
grfi=getModelInfo("rf")
grfi$rf$parameters
rf.grid <- expand.grid(mtry=1:10) # number of variables available for splitting at each tree node

# XGboost
gxbi=getModelInfo("xgbTree")
gxbi$xgbTree$parameters
gbm.grid <-  expand.grid(max_depth = c(3, 5, 7), 
                        nrounds = (1:10)*50,    # number of trees
                        # default values below
                        eta = 0.3,
                        gamma = 0,
                        subsample = 1,
                        min_child_weight = 1,
                        colsample_bytree = 0.6)

# neuralnet
gnni$neuralnet$parameters
neunet.grid= expand.grid(layer1 = 10, layer2 = 10, layer3 = 10)

# nnet
grfi=getModelInfo("nnet")
grfi$nnet$parameters
nnet.grid = expand.grid(size = seq(from = 2, to = 10, by = 2), # number of neurons units in the hidden layer 
                        decay = seq(from = 0.1, to = 0.5, by = 0.1)) # regularization parameter to avoid over-fitting 
```

Train models
```{r include = TRUE, message=F, warning=F, eval=FALSE}
# Random forest
rf_model <- caret::train(x = train[,1:(ncol(train)-1)], y = train$veg,
                         method = "rf", metric="Accuracy", trainControl = tc, tuneGrid = rf.grid)
rf_model
plot(rf_model)

# XG boost
xg_model = train(veg~., data = train, method = "xgbTree", metric="Accuracy", 
                 trainControl = tc, tuneGrid = gbm.grid)
xg_model
plot(xg_model)

# nnet
nnet_model <- caret::train(x = trainx[,(1:ncol(trainx)-1)], y = trainx$veg, 
                           method = "nnet", metric="Accuracy", 
                           trainControl = tc, tuneGrid = nnet.grid)
nnet_model
plot(nnet_model)
```

Produce performance metrics
Use model to make predictions on test data
```{r include = TRUE, message=F, warning=F, eval=FALSE}
pred_rf = predict(rf_model, test)
pred_xg = predict(xg_model, test)
#pred_neunet = predict(neunet_model, testx)
pred_nnet = predict(nnet_model, testx)
```

De-standardize predictions based on the mean and sd obtained from training data
```{r include = TRUE, message=F, warning=F, eval=FALSE}
#pred_neunet=pred_neunet*sd(train$agb)+mean(train$agb)
#pred_nnet=pred_nnet*sd(train$agb)+mean(train$agb)
```

Calculate accuracy estimates by comparing observed vs predicted values for response variable
```{r include = TRUE, message=F, warning=F, eval=FALSE}
test_y = test[, ncol(test)] 

# rf
confusionMatrix(pred_rf, test$veg)

# xgboost
confusionMatrix(pred_xg, test$veg)

# nnet
confusionMatrix(pred_nnet, test$veg)
```

### Make spatial predictions
```{r include = TRUE, message=F, warning=F, eval=FALSE}
# Upload datastack
kgz=rast(paste0(sourcedir, "/RasterStack_KGZ.tiff"))
names(kgz) = c("Aspect", "B2_Ref", "B3_Ref", "B4_Ref", "B5_Ref",
               "B6_Ref", "B7_Ref", "DEM", "SettlementDistance", 
               "LS_Factor", "NDVI", "NDWI", "Sky_View_Factor",
               "Slope", "Topographic_Wetness_Index", "Valley_Depth") 

plot(kgz[[1]])
e=draw("extent")
kgzrz=crop(kgz, e)
plot(kgzrz[[1]])
kgzrzdf=as.matrix(kgzrz, mode="numeric")
kgzrzdf=subset(kgzrzdf, complete.cases(kgzrzdf))
colnames(kgzrzdf)==names(train)

# Apply the models to the spatial data
# Let's create a reference raster template
refrast=sum(kgzrz)
refrast[!is.na(refrast)] <- 1

# Random forest
rf_prediction=predict(rf_model, kgzrzdf)
rf_rast=refrast
rf_rast[!is.na(rf_rast)] <- rf_prediction
plot(rf_rast)
writeRaster(rf_rast, "rf_rast_c.tif")

#  xg_boost
xgb_prediction=predict(xg_model, kgzrzdf)
xgb_rast=refrast
xgb_rast[!is.na(xgb_rast)] <- xgb_prediction
plot(xgb_rast)
writeRaster(xgb_rast, "xgb_rast_c.tif")

#  nnet
kgrzdfx=data.frame(apply(kgzrzdf, 2, FUN=zscore))
nnet_prediction = predict(nnet_model,kgrzdfx)
# Values need to be rescaled again
nnet_rast=refrast
nnet_rast[!is.na(nnet_rast)] <- nnet_prediction
plot(nnet_rast)
writeRaster(nnet_rast, "nnet_rast_c.tif")
```

### Lab report

This week's lab will be an expansion of the lab from previous weeks. Therefore it will count as the report from last and this week's lab (8 points in total).

Adapt the code illustrated today to identify suitable values for hyperparameters associated with the application of the four models for the data in your final project.

Based on your results, report the optimal parameters for both algorithms based on a validation analysis and add a figure that supports your selection.

Perform an accuracy assessment for both algorithms using the test set and report your results. Based on your results, explain what algorithm produces the best performance.