---
title: 'Applied Machine Learning for Spatial Analysis. Lab Week 8: Supervised classification'
author: "Victor Gutierrez (victorhugo@temple.edu)"
output:
  html_document:
    df_print: paged
  pdf_document: default
  word_document: default
---

```{r include = FALSE}
# Load screenshots
wd="/Users/tug61163/Library/CloudStorage/OneDrive-TempleUniversity/Repositories/Gitrepo/MALESA_Code"
setwd(wd)
library(knitr)    # For knitting document and include_graphics function
library(ggplot2)  # For plotting
library(png)      # For grabbing the dimensions of png files

fig1path=paste(wd, "Fig1.png", sep="/")
#fig2path=paste(wd, "Sel2Edit.png", sep="/")
#fig3path=paste(wd, "Triangle.png", sep="/")
#fig4path=paste(wd, "Fig4.png", sep="/")
#fig1 = readPNG(fig1path)
#fig2 = readPNG(fig2path)
#fig3 = readPNG(fig3path)
#fig4 = readPNG(fig4path)
```

## Goals
To learn how to fit a supervised classification and then use it to produce predictions using different machine learning algorithms.

## Total score
The lab counts for up to 4 points towards the final grade of the course.

## Lab instructions
1.	Launch R Studio and open a new R script: File/ New File/ R Script. Then save it as a new file: File/ Save As…
2.	Read the instructions below step by step. Copy and paste each chunk of code at a time in your R script. Select the code with your mouse or shift/arrow keys and then run it by pressing the keys control-enter simultaneously.

## Lab overview
For this lab, we are going to use the mosquito data that we 

Have fun!

### 1. System setup
```{r include = TRUE, message=F, warning=F, eval=FALSE}
wd=("/Users/tug61163/Library/CloudStorage/OneDrive-TempleUniversity/Courses/MALESA/2022Fall/Week9")
setwd(wd)
library(terra)
library(sf)
library(ggplot2)
library(DataExplorer)
library(randomForest)
library(gbm)
library(e1071)
library(rpart)
library(rattle)
```

#### Load spatial files
```{r include = TRUE, message=F, warning=F, eval=FALSE}
kgz=rast("RasterStack_KGZ.tiff")
names(kgz) = c("Aspect", "B2_Ref", "B3_Ref", "B4_Ref", "B5_Ref",
               "B6_Ref", "B7_Ref", "DEM", "SettlementDistance", 
               "LS_Factor", "NDVI", "NDWI", "Sky_View_Factor",
               "Slope", "Topographic_Wetness_Index", "Valley_Depth") 

classes=st_read("Vegetation Plots.shp")
classes=vect(classes)
```

#### Extract data to dataframe for modeling
```{r include = TRUE, message=F, warning=F, eval=FALSE}
sampdata=terra::extract(kgz, classes, xy=TRUE)
sampdata$veg=classes$Vegetation
sampdata=subset(sampdata, complete.cases(sampdata))
sampdata$veg=as.factor(sampdata$veg)
```

#### Explore the data
```{r include = TRUE, message=F, warning=F, eval=FALSE}
plot_bar(sampdata)

pairs(sampdata[,c(2:17)], upper.panel = NULL,
      diag.panel = panel.hist)

ggplot(sampdata, aes(x = veg, y = DEM)) + geom_boxplot()
ggplot(sampdata, aes(x = veg, y = NDWI)) + geom_boxplot()
ggplot(sampdata, aes(x = veg, y = B5_Ref)) + geom_boxplot()
ggplot(sampdata, aes(x = veg, y = Slope)) + geom_boxplot()
```

#### Data preparation
Subset data before modeling to remove irrelevant variables and also to make the variables in the dataframe match the variables in the raster stack
```{r include = TRUE, message=F, warning=F, eval=FALSE}
sampdata=sampdata[,-c(1, 18:19)]
```

Normalize your data. This is imperative for some ML methods such as SVM and kNN. From lab 5
```{r include = TRUE, message=F, warning=F, eval=FALSE}
zscore=function(x){
  zval=(x-mean(x))/sd(x)
  return(zval)
}

for(xval in 1:(ncol(sampdata)-1)){
  sampdata[,xval]=zscore(sampdata[,xval])
}
```

#### Fit different ML methods

##### Decission trees
```{r include = TRUE, message=F, warning=F, eval=FALSE}
dt=rpart(veg~., data=sampdata, method="class")
# Check the arguments for rpart.control
# Explain the tree outputs

#printcp(dt) #displays the results
#plotcp(dt)
#summary(dt)
dt$variable.importance
plot(dt, uniform=TRUE, 
     main="Classification Tree for plants")
text(dt, use.n=TRUE, all=TRUE, cex=.8)
fancyRpartPlot(dt, caption = NULL)

#Pruning the tree
dt$cptable
printcp(dt)
plotcp(dt)
dtprune=prune(dt, cp= dt$cptable[which.min(dt$cptable[,"xerror"]),"CP"])

# plot the pruned tree 
plot(dtprune, uniform=TRUE, 
     main="Pruned Classification Tree for Kyphosis")
text(dtprune, use.n=TRUE, all=TRUE, cex=.8)
```

##### Random forest
```{r include = TRUE, message=F, warning=F, eval=FALSE}
rf=randomForest(veg~., data=sampdata, ntree=200, importance=TRUE)
rf
rf$importance
```

##### Gradient boosting machine 
```{r include = TRUE, message=F, warning=F, eval=FALSE}
gb <- gbm(
  formula = veg ~ .,
  distribution = "multinomial",
  data = sampdata,
  n.trees = 10000,
  interaction.depth = 1,
  shrinkage = 0.001,
  cv.folds = 5,
  n.cores = NULL, # will use all cores by default
  verbose = FALSE
)  

# You can see that the multinomial option is broken
# Let's convert it to a binomial prediction by predicting just one vegetation community
sampdata$AcoPra=rep(NA, nrow(sampdata))
sampdata$AcoPra[sampdata$veg=="Aco-Pra"]=1
sampdata$AcoPra[sampdata$veg!="Aco-Pra"]=0
sampsub=sampdata[,-17]
gb <- gbm(
  formula = AcoPra ~ .,
  distribution = "bernoulli",
  data = sampsub,
  n.trees = 10000,
  interaction.depth = 1,
  shrinkage = 0.001,
  cv.folds = 5,
  n.cores = NULL, # will use all cores by default
  verbose = FALSE
)  

print(gb)
gbm.perf(gb, method = "cv") # Optimal number of iterations

# Variable importance
summary(
  gb, 
  cBars = 10,
  method = relative.influence, # also can use permutation.test.gbm
  las = 2
)

```

##### Support vector machine
```{r include = TRUE, message=F, warning=F, eval=FALSE}
sv = svm(veg~., data = sampdata, type = 'C-classification', kernel = "sigmoid", cost = 10, scale = FALSE)
print(sv)
plot(sv, data=sampdata, formula=NDWI~DEM)

# Fit Support Vector Machine model to data set
svtune <- tune(svm, veg~., data = sampdata, kernel = "sigmoid",
                 ranges = list(cost = c(0.001, 0.01, 0.1, 1, 5, 10, 100)))
# extract the best model
bestmod <- svtune$best.model
plot(bestmod, data=sampdata, formula=NDWI~DEM)

# Let's try with a different type of kernel
sv2 = svm(veg~., data = sampdata, type = 'C-classification', kernel = "radial", cost = 10, scale = FALSE)
print(sv2)
plot(sv2, data=sampdata, formula=NDWI~DEM)

svtune2 <- tune(svm, veg~., data = sampdata, kernel = "radial",
               ranges = list(cost = c(0.001, 0.01, 0.1, 1, 5, 10, 100)))
bestmod2 <- svtune2$best.model
plot(bestmod2, data=sampdata, formula=NDWI~DEM)
```

#### Produce predictions and save them
```{r include = TRUE, message=F, warning=F, eval=FALSE}
dtpred=terra::predict(kgz, dt)
plot(dtpred)

rfpred=terra::predict(kgz, rf)
plot(rfpred)

gbpred=terra::predict(kgz, gb, type="response")
plot(gbpred)

# To predict svm
kgz.df <- terra::values(kgz)
names(kgz.df) <-names(kgz)
svm.pred <- predict(sv, kgz.df)

svpred=terra::predict(kgz, sv)
plot(svpred)

writeRaster(dtpred, "dtpred.tif")
writeRaster(rfpred, "rfpred.tif")
writeRaster(gbpred, "gbpred.tif")
writeRaster(svpred, "svpred.tif")
```

#### Lab report
Instructions for this lab report will be provided next week, after overviewing supervised regression