---
title: 'Applied Machine Learning for Spatial Analysis. Lab Week 2: R environment and objects'
author: "Victor Gutierrez (victorhugo@temple.edu)"
output:
  html_document:
    df_print: paged
  pdf_document: default
  word_document: default
---

```{r include = FALSE}
# Load screenshots
wd="/Users/tug61163/Documents/Courses/MALESA/2022Fall/Class1/"
library(knitr)    # For knitting document and include_graphics function
library(ggplot2)  # For plotting
library(png)      # For grabbing the dimensions of png files
setwd(wd)
fig1path=paste(wd, "Fig1.png", sep="/")
#fig2path=paste(wd, "Sel2Edit.png", sep="/")
#fig3path=paste(wd, "Triangle.png", sep="/")
#fig4path=paste(wd, "Fig4.png", sep="/")
#fig1 = readPNG(fig1path)
#fig2 = readPNG(fig2path)
#fig3 = readPNG(fig3path)
#fig4 = readPNG(fig4path)
```
## Lab due
Sep 12 2022

## Goal
To learn how to download and pre-process spatial data for machine learning application.

## Total score
The lab counts for up to 4 points towards the final grade of the course.


## Lab instructions
1.	Launch R Studio and open a new R script: File/ New File/ R Script. Then save it as a new file: File/ Save As…
2.	Read the instructions below step by step. Copy and paste each chunk of code at a time in your R script. Select the code with your mouse or shift/arrow keys and then run it by pressing the keys control-enter simultaneously.

I strongly recommend this introductory R book: https://cran.r-project.org/doc/manuals/r-release/R-intro.pdf 

Have fun!
1. Define environment and load required libraries
```{r include = TRUE, message=F, warning=F, eval=FALSE}
wd="/Users/tug61163/Documents/Courses/MALESA/2022Fall/Class2/Lab2"
setwd(wd)
library(sf)
library(terra)
library(osmdata)
```

12. Download osm data. Check all the available features in osm data
```{r include = TRUE, message=F, warning=F, eval=FALSE}
available_features()
```

You can also check all the tags within each feature
```{r include = TRUE, message=F, warning=F, eval=FALSE}
available_tags("highway")
```

Let's first create a bounding box of the extent covered by our point file
```{r include = TRUE, message=F, warning=F, eval=FALSE}
emat=t(matrix(e, 2, 2))
ematsub=t(matrix(c(-82.82822,-81, 37.66454 , 40),2,2))
rownames(ematsub)=c("x","y")
colnames(ematsub)=c("min", "max")
ematsub

ways<- ematsub %>%
  opq() %>%
  add_osm_feature(key = "highway") %>%
  osmdata_sf()

ways$osm_lines

ggplot() +
  geom_sf(data =ways$osm_lines)
```

Another option is to download the file from http://download.geofabrik.de 
```{r include = TRUE, message=F, warning=F, eval=FALSE}
download.file("http://download.geofabrik.de/north-america/us/new-york-latest-free.shp.zip", paste(wd, "new-york-latest-free.shp.zip", sep="/"))

download.file("http://download.geofabrik.de/north-america/us/pennsylvania-latest-free.shp.zip", paste(wd, "pennsylvania-free.shp.zip", sep="/"))

dir.create(paste(wd, "NY", sep="/"))
dir.create(paste(wd, "PA", sep="/"))
unzip("new-york-latest-free.shp.zip",exdir= paste(wd, "NY", sep="/") )
unzip("pennsylvania-free.shp.zip", exdir= paste(wd, "PA", sep="/") )
```

13.Open the roads file
```{r include = TRUE, message=F, warning=F, eval=FALSE}
NYroads=vect(paste(paste(wd, "NY", sep="/"), "gis_osm_roads_free_1.shp", sep="/"))
PAroads=vect(paste(paste(wd, "PA", sep="/"), "gis_osm_roads_free_1.shp", sep="/"))
```

14. Plot everything
```{r include = TRUE, message=F, warning=F, eval=FALSE}
plot(elev)
plot(pm25tSub, add=TRUE)
plot(NYroads, add=TRUE)
plot(PAroads, add=TRUE)
```

```{r include = TRUE, message=F, warning=F, eval=FALSE}
```

```{r include = TRUE, message=F, warning=F, eval=FALSE}
```

```{r include = TRUE, message=F, warning=F, eval=FALSE}
```

```{r include = TRUE, message=F, warning=F, eval=FALSE}
```

```{r include = TRUE, message=F, warning=F, eval=FALSE}
```