---
title: 'Applied Machine Learning for Spatial Analysis. Lab Week 4: Data Preparation'
author: "Victor Gutierrez (victorhugo@temple.edu)"
output:
  html_document:
    df_print: paged
  pdf_document: default
  word_document: default
---

```{r include = FALSE}
# Load screenshots
wd="/Users/tug61163/Library/CloudStorage/OneDrive-TempleUniversity/Repositories/Gitrepo/MALESA_Code"
setwd(wd)
library(knitr)    # For knitting document and include_graphics function
library(ggplot2)  # For plotting
library(png)      # For grabbing the dimensions of png files

fig1path=paste(wd, "Fig1.png", sep="/")
#fig2path=paste(wd, "Sel2Edit.png", sep="/")
#fig3path=paste(wd, "Triangle.png", sep="/")
#fig4path=paste(wd, "Fig4.png", sep="/")
#fig1 = readPNG(fig1path)
#fig2 = readPNG(fig2path)
#fig3 = readPNG(fig3path)
#fig4 = readPNG(fig4path)
```
## Lab due
October 03

## Goal
To learn how to perform exploratory data analysis.

## Total score
The lab counts for up to 4 points towards the final grade of the course.

## Lab instructions
1.	Launch R Studio and open a new R script: File/ New File/ R Script. Then save it as a new file: File/ Save As…
2.	Read the instructions below step by step. Copy and paste each chunk of code at a time in your R script. Select the code with your mouse or shift/arrow keys and then run it by pressing the keys control-enter simultaneously.

## Lab overview
For this lab, we are going to use the mosquito data that we 

Have fun!

### 1. System setup
```{r include = TRUE, message=F, warning=F, eval=FALSE}
wd=("/Users/tug61163/Library/CloudStorage/OneDrive-TempleUniversity/Courses/MALESA/2022Fall/Week6")
setwd(wd)
library(terra)
library(sf)
library(DataExplorer)
library(GGally)
library(dplyr)
```

#### Load files
```{r include = TRUE, message=F, warning=F, eval=FALSE}
load("alldata.RData")
```

#### Preparing the data
Before exploring the data, let's create a binary column representing mosquito presences/absences.
```{r include = TRUE, message=F, warning=F, eval=FALSE}
alldata$PRESENCE=alldata$ADULTS
alldata$PRESENCE[alldata$PRESENCE>0]=1
alldata$PRESENCE=as.factor(alldata$PRESENCE)
#alldata=
```

#### Identify the structure of the data
```{r include = TRUE, message=F, warning=F, eval=FALSE}
str(alldata)
plot_str(alldata)
summary(alldata)
head(alldata)
```

#### Identify and remove observations with missing values and data duplicates
```{r include = TRUE, message=F, warning=F, eval=FALSE}
nrow(alldata)
alldata=subset(alldata, complete.cases(alldata))
nrow(alldata)

# Identify and remove duplicates
dups=duplicated(alldata)
alldata[dups,]
nrow(alldata)
alldata=alldata[!dups,]
nrow(alldata)
```

#### Explore the characteristics of individual variables
```{r include = TRUE, message=F, warning=F, eval=FALSE}
names(alldata)
datasubset=alldata[,c(2, 6:10, 13:33)]
plot_bar(datasubset)
plot_qq(datasubset)
plot_density(datasubset)
plot_prcomp(datasubset[,c(2,3, 7:24)], nrow=2L, ncol=2L)

```

#### Check interactions between pairs of data 
```{r include = TRUE, message=F, warning=F, eval=FALSE}
cormat=cor(datasubset[,c(2, 3, 7:20, 22:26)], use="complete.obs", method="pearson")
plot_correlation(datasubset)

pairs(datasubset[,c(7:13)])

panel.hist <- function(x, ...) {
    usr <- par("usr")
    on.exit(par(usr))
    par(usr = c(usr[1:2], 0, 1.5))
    his <- hist(x, plot = FALSE)
    breaks <- his$breaks
    nB <- length(breaks)
    y <- his$counts
    y <- y/max(y)
    rect(breaks[-nB], 0, breaks[-1], y, col = rgb(0, 1, 1, alpha = 0.5), ...)
    # lines(density(x), col = 2, lwd = 2) # Uncomment to add density lines
}
pairs(datasubset[,c(7:13)], upper.panel = NULL,
      diag.panel = panel.hist)
plot_bar(datasubset, by="PRESENCE")
```

#### Plot data interactions per groups
```{r include = TRUE, message=F, warning=F, eval=FALSE}
options(repr.plot.width = 20, repr.plot.height = 10)
alldata %>% 
  select("roughness", "tmea", "income", "latino", "TRI") %>%
  ggpairs(mapping = aes(color = alldata$PRESENCE, alpha = 0.5))

boxplot(income~PRESENCE,data=alldata, main="income per mosquito presence", 
   xlab="presence/absence", ylab="income")

boxplot(TRI~PRESENCE,data=alldata, main="TRI per mosquito presence", 
   xlab="presence/absence", ylab="TRI")

pairs(datasubset[,c(7:13)],                     # Data frame of variables
      labels = colnames(datasubset[,c(7:13)]),  # Variable names
      pch = 21,                                 # Pch symbol
      bg = rainbow(3)[datasubset$PRESENCE],     # Background color of the symbol (pch 21 to 25)
      col = rainbow(3)[datasubset$PRESENCE],    # Border color of the symbol
      main = "Iris dataset",                    # Title of the plot
      row1attop = TRUE,                         # If FALSE, changes the direction of the diagonal
      gap = 1,                                  # Distance between subplots
      cex.labels = NULL,                        # Size of the diagonal text
      font.labels = 1)                          # Font style of the diagonal text
```

#### Produce report
Let's produce a report using the package "DataExplorer"
```{r include = TRUE, message=F, warning=F, eval=FALSE}
alldata %>%
    create_report(
        output_file = paste("Report", format(Sys.time(), "%Y-%m-%d %H:%M:%S %Z"), sep=" - "),
        report_title = "EDA Report - Mosquito presence",
        y = "PRESENCE"
    )
```

#### Assess and mitigate spatial bias
```{r include = TRUE, message=F, warning=F, eval=FALSE}
# Upload reference raster
load("alldata.RData")
datastack=rast("datastack.tif")
SpatData=vect("SpatData.shp")
names(datastack)=names(alldata[13:32])

# create a SpatRaster with the extent of the raster dataset
r <- datastack[[1]]
# set the resolution of the cells to fifty times the original resolution
res(r) <- 50*(res(r))
# extend (expand) the extent of the SpatRaster a little
r <- extend(r, ext(r)+1)
# sample:
set.seed(13)
samp <- spatSample(SpatData, size=1, "random", strata=r)
# to illustrate the method and show the result
p <- as.polygons(r)
plot(p, border='gray')
points(SpatData)
# selected points in red
points(samp, cex=1, col='red', pch='x')

# Extract values of variables for the random subset
datasamp=extract(datastack, samp)
nrow(datasamp)
nrow(samp)
subdata=cbind(samp, datasamp)
save(subdata, file="subdata.RData")

# Compare the distribution of the values of the dataset before and after filtering with the distribution that would be obtained under an ideal sampling design, in this case it would be a random sample

# Before filtering
reforig=spatSample(datastack[[1]], size=nrow(SpatData), "random")
cdf1=ecdf(alldata$ppt_)
cdf2=ecdf(reforig$ppt_)
plot(cdf1)
plot(cdf2, col="red", add=T)

# After filtering
reffilt=spatSample(datastack[[1]], size=nrow(samp), "random")
cdf1=ecdf(subdata$ppt_)
cdf2=ecdf(reffilt$ppt_)
plot(cdf1)
plot(cdf2, col="red", add=T)
```

It looks like it hasn't changed that much. We will have to constraint the geographic domain fo the data to include only the samples within urban areas. For that purpose, we will download a map representing urban areas only from the US census: https://www.census.gov/cgi-bin/geo/shapefiles/index.php 

Then, we will use the map to mask out any data that is not within urban areas
```{r include = TRUE, message=F, warning=F, eval=FALSE}
urban=vect("tl_2015_us_uac10.shp")
urban=project(urban, datastack)
urban=crop(urban, ext(datastack))
writeVector(urban, "urban.shp")
urbanrast=rasterize(urban, datastack)
datastackurb=mask(datastack, urbanrast)
plot(datastackurb[[1]])
```

Let's repeat the procedure with only the masked areas
```{r include = TRUE, message=F, warning=F, eval=FALSE}
# Select points within urban polygons
urbansf=read_sf("urban.shp")
SpatDataSf=read_sf("SpatData.shp")
ptsInUrban=st_intersection(SpatDataSf, urbansf)
ptsInUrban=vect(ptsInUrban)
plot(urbanrast)
plot(ptsInUrban, add=T)
ptsInUrban=ptsInUrban[,9]



# create a SpatRaster with the extent of the raster dataset
r <- datastackurb[[1]]
# set the resolution of the cells to fifty times the original resolution
res(r) <- 50*(res(r))
# extend (expand) the extent of the SpatRaster a little
r <- extend(r, ext(r)+1)
# sample:
set.seed(13)
samp <-spatSample(ptsInUrban, size=1, "random", strata=r)
# to illustrate the method and show the result
p <- as.polygons(r)
plot(p, border='gray')
points(ptsInUrban)
# selected points in red
points(samp, cex=1, col='red', pch='x')
```

## Lab deliverables (due on Sep 19)

1. 







```{r include = TRUE, message=F, warning=F, eval=FALSE}
```

```{r include = TRUE, message=F, warning=F, eval=FALSE}
```





