---
title: 'Remote Sensing Lab 10: Unsupervised and Supervised classification'
author: "Victor Gutierrez (victorhugo@temple.edu)"
output:
  html_document:
    df_print: paged
  pdf_document: default
  word_document: default
always_allow_html: yes
---

```{r include = FALSE}
# Load screenshots
wd="/Users/tug61163/Library/CloudStorage/OneDrive-TempleUniversity/Courses/IntroRemoteSensing/2023Fall/Class10_Classification/Lab10Materials"
setwd(wd)
library(knitr)    # For knitting document and include_graphics function
```

## Lab due
November 06, 2024

## Goals
1. To learn graphical approaches to assess class separability for land cover classification.
2. To learn methods for unsupervised and supervised image classification and their implementation in R.

## Total score
The lab counts for up to 4 points towards the final grade of the course.

## Lab instructions
This lab assumes that some of the procedures learned in previous labs for downloading and pre-processing satellite images in R as well as deriving band transformations and spectral indices are already mastered and therefore are not covered here. Please refer to previous labs to implement such previous steps.

1. Open R Studio and then a new R script. Copy and paste each chunk of the code below in your new R script and run it trying to understand the purpose, logic and syntaxis of each line. Make sure the code runs with no errors before moving to the next one.

2.	Answer the questions in the answer sheet and submit your along with the required files to canvas.

Have fun!

## Lab overview
For this lab we will perform an unsupervised classification using the k-means method. Then we will...

### A. Setup environment, and load required libraries and functions. 

#### 1. Setup environment, and load required libraries
Use the function install.packages("") to install any unavailable libraries
```{r include = TRUE, message=F, warning=F, eval=FALSE}
wd="/Users/tug61163/Library/CloudStorage/OneDrive-TempleUniversity/Courses/IntroRemoteSensing/2024Fall/Class10_Class"
#wd="/Users/tug61163/Downloads/OneDrive_1_10-10-2023"
setwd(wd)

library(terra)
library(sf)
library(RStoolbox)
library(RColorBrewer)
library(rgl)
#library(randomForest)

#library(prettymapr)
# library(raster)
# library(rgdal)
# library(RStoolbox)

# library(stringr)

# Load this entire function that I created. We will use it later to visualize and interpret some data
plotSpectra=function(dataset=outdata, bandnames=c("B3_dn", "B4_dn", "B5_dn"),
                     classfield=1, classlabels= sort(unique(dataset[,classfield])),
                     classcol=sample(colors(), size=length(classlabels))){
  require(rgl)
  #Assign colors to different labels
  dataset=dataset[order(dataset[[classfield]]), ]
  cols=rep(NA, length(classlabels))
  for (i in 1:length(classlabels)){
    cols[which(dataset[,classfield]==classlabels[i])]=classcol[i]
  }
  
  return(plot3d(dataset[,as.character(bandnames[1])],
                dataset[, as.character(bandnames[2])], 
                dataset[, as.character(bandnames[3])],
                xlab=bandnames[1],
                ylab=bandnames[2],
                zlab=bandnames[3],
                col=cols)) 
  legend3d("topleft", legend = classlabels, 
           pch = 16, col = classcol  , cex=1, inset=c(0.02))
}
```

#### 2. Load the data
Open the images that you produced in the previous lab. These will be used as input variables for the classification.
```{r include = TRUE, message=F, warning=F, eval=FALSE}
dir() # check that the files you need are in the directory
L8name=rast("L8rsz.tif")
#L8indices=rast("L8indices.tif")
tassCap=rast("L8tascap.tif")
tcnames=c("brightness", "greenness", "wetness" )
names(tassCap)=tcnames

# Upload the shapefile representing the training polygons that you collected for different land cover classes in the previous lab.
```{r include = TRUE, message=F, warning=F, eval=FALSE}
classes=st_read("Training_Pucallpa_2015_Sel.shp") 

# make sure the coordinate system is the same for the raster and the vector files
crs(tassCap) == crs(classes)
classes=st_transform(classes, crs(tassCap))
crs(tassCap) == crs(classes)

unique(classes$CLASS_NAME)

#Notice that there are rows with NAs in the name that we have to eliminate
classes <- subset(classes, !is.na(CLASS_NAME))
unique(classes$CLASS_NAME)
#classesV=vect(classes)
```

#### 3.  Rescale the data
K-means tends to create oblong objects and so it is not very good at handling data that comes at different scales. Therefore it is strongly adviced to re-scale all data.

A common method to rescale the data is by transforming them into zscores. From basic statstics, a zscore represents how many standard deviations a single value is away from the mean. Therefore a zscore has a mean of zero and a standard deviation of one.

```{r include = TRUE, message=F, warning=F, eval=FALSE}
tassCapZ=scale(tassCap)
tassCapZ
```

### B.  Perform an unsupervised classification. 
As the number of classes, select the same number as the number classes considered in the collection of training polygons.

If you receive this message "did not converge in 50 iterations ", it means that the algorithm didn't converge to a maximum. You still will probably obtain sensitive results but you might want to increase the parameter iter.max (eg. iter.max=100) and assess whether that improves the classification.

```{r include = TRUE, message=F, warning=F, eval=FALSE}
unsupervised=unsuperClass(tassCapZ, nClasses=13)
plot(unsupervised$map, col=sample(colors(), 13))
```

### C. Perform supervised classification

#### 1. Extract the pixel values for each band in a dataframe
```{r include = TRUE, message=F, warning=F, eval=FALSE}
sampdata=terra::extract(tassCap, vect(classes), xy=TRUE)

# create a new row in the data frame to store the names of the land covers for the pixels in each polygon
sampdata$LC=rep(NA, nrow(sampdata))
for (i in 1: nrow(classes)){
  sampdata$LC[which(sampdata$ID==i)]=rep(classes$CLASS_NAME[i], length(sampdata$LC[which(sampdata$ID==i)]))
}

# Remove any incomplete observations
sampdata=subset(sampdata, complete.cases(sampdata))

# reformat the colummn containing land cover classes as factor so that they are considered as categories
sampdata$LC=as.factor(sampdata$LC)
```

#### 2. Visualize spectral space
Produce the 3d plot by implementing the plotSpectra function that you loaded earlier.

NOTE: If you are working with a Mac computer, you will need to have the XQuartz software installed in your computer. If it is not installed, you can download it from here: https://www.xquartz.org/

```{r include = TRUE, message=F, warning=F, eval=FALSE}

length(unique(sampdata$LC))

# display color palettes and select one for plotting each land cover class
display.brewer.all()
display.brewer.pal(12, "Paired")

# Since I have 13 classes and the palette has 12 colors, let's add one more color 
classcolors= c(brewer.pal(n = 12, name = "Paired"),"gray")
names(sampdata) 

plotSpectra(dataset=sampdata, bandnames=tcnames, classfield=7,
            classlabels=sort(unique(sampdata$LC)),
            classcol=classcolors)

# Make sure that the name in the legend coincides with the name of the column representing class names. In the example below is Class.ID. You can change the size of the legend by modifying the value in the argument cex
legend3d("topleft", legend= sort(unique(sampdata$LC)),
         col=classcolors, pch=3, cex=.4, inset=c(0.02))
```

#### 3. Produce a supervised classification
```{r include = TRUE, message=F, warning=F, eval=FALSE}
supervised=superClass(tassCap, classes, trainPartition=.7,
                      responseCol = "CLASS_NAME" )

# retrieve variable importance from the supervised object
supervised$model$finalModel$importance

# plot the map
plot(supervised$map, col=classcolors)
```

#### 4.  Save results
```{r include = TRUE, message=F, warning=F, eval=FALSE}
save(unsupervised, file="unsupervised.RData")
writeRaster(unsupervised$map, "unsupervised.tif")

save(supervised, file="supervised.RData")
writeRaster(supervised$map, "supervised.tif")
```

### C. Map comparison
Open both classification results in QGIS or any other GIS software. Use a high resolution base map to interpret the most likely land cover category for each cluster in the unsupervised classification map. Enter the names of the interpreted classes below in the "catnamesUnsup" object. Plot maps and figures including a legend, scale bar and north arrow and save them as a single pdf document
```{r include = TRUE, message=F, warning=F, eval=FALSE}
#unsupercol=sample(colors(), 13) # These are the colors to be used for mapping the unsupervised map
plot(unsupervised$map, col=classcolors, axes=FALSE, legend=T, box=FALSE)

# Interpret the names of the classes in the unsupervised map and assign them to each category number below. You should do this through visual interpretation of the different colors, using QGIS or another GIS software
catnamesUnsup=c("Adolescent Oil Palm", "Bare Ground", "Lake", "Mature Oil Palm", "New Burn Scar", "Old Burn Scar", "Old Growth Forest", "Pasture",  "River", "Sand Banks", "Secondary Forest", "Urban", "Young Oil Palm")

# Assign class names to the class numbers classified through the unsupervised and supervised classification
unsuperclass_levels=data.frame(cbind(seq(1:length(catnamesUnsup)), catnamesUnsup))
unsuperclass_levels$V1=as.numeric(unsuperclass_levels$V1)
levels(unsupervised$map)=unsuperclass_levels

superclass_levels=data.frame(cbind(seq(1:length(colnames(supervised$modelFit[[2]]$table))), colnames(supervised$modelFit[[2]]$table)))
superclass_levels$X1= as.numeric(superclass_levels$X1)
levels(supervised$map)=superclass_levels

pdf("VGutierrezLab10.pdf", paper="USr", width=15)

# PLOT THE UNSUPERVISED CLASSIFICATION
# assign intuitive colors to the different classes. In this case, I will use the same colors as the supervised classification
plot(unsupervised$map, col=classcolors)
# Add north arrow, scalebar.

# PLOT THE SUPERVISED CLASSIFICATION
plot(supervised$map, col=classcolors)

dev.off()
```

## Lab deliverables

1.	Add three X, Y, Z figures from different angles of the spectral space obtained through the application of the “plotSpectra()” function with the three bands that are best to discriminate between the land cover categories selected. Make sure the axes are at an angle that optimizes the visualization of the contrast in pixel values between categories (0.5 pts).

2. Respond the questions below based on the visualization of the spectral space (0.3 pts):
a. Which are the two land covers that you expect to be classified the most accurately? ___________________
b. Which are the two land covers that you expect to be the most confused? _________________

3. Upload in canvas a single pdf file with the map obtained from the unsupervised and supervised classifications.  (0.2 pts).

4. Upload to canvas a single word document showing three zoomed-in images for one polygon per each land cover class. The first image corresponds to an RGB color composite from the original satellite image, the second one will be the results of the unsupervised classification and the third will be the results of the supervised classification. All the zoomed in images should include the outline of the polygon with no fill. You can produce these figures in QGIS or any other GIS software (1.2 pts).

5. Based on your visual interpretation of the classification results, please indicate which type of classification (unsupervised vs supervised) represents best the land covers of interest (0.2 pts)____________ . 

6. For the best classification results, please respond the questions below (0.6 pts):
a. Which two land covers are best represented in your best classification (either supervised or unsupervised)? 
b. Which two land covers have the worst representation?
c. Do these results agree with the confusion that you identified based on the analysis of the spectral space (point 2)? 


